Visuomotor mismatch EEG responses in occipital cortex of freely moving human subjects

Version 1

Abstract

Likely the strongest predictor of visual feedback is self-motion. In mice, the coupling between movement and visual feedback is learned with first visual experience of the world (Attinger et al., 2017), and brief perturbations of the coupling result in strong visuomotor mismatch responses in visual cortex that possibly reflect prediction errors (Keller et al., 2012; Zmarz and Keller, 2016). In humans, predictive coding has primarily been studied using oddball paradigms which rely on violations of stimulus probability based on recent sensory history. It was still unclear, however, whether humans exhibit visuomotor mismatch responses similar to those observed in mice. This question was important for two reasons. First, visuomotor mismatch responses in humans constitute a basis to start translating the mechanistic understanding of the circuit that computes these responses from mouse to human cortex. Second, a paradigm that can trigger strong prediction error responses and consequently requires shorter recording times would simplify experiments in a clinical setting. Here, by combining a wireless EEG recording system with virtual reality headset, we found robust visuomotor mismatch responses in human cortex that were characterized by a reversed polarity relative to visual evoked responses and a greater signal power than both visual responses and oddball mismatch responses.

Introduction

Predictive processing is a framework for understanding brain function. It proposes that the brain constructs internal models of the world based on regularities in past sensory input and sensorimotor loops to predict upcoming sensory input. The brain does this by minimizing the mismatches between predicted and actual sensory input. These mismatches, known as prediction errors, play a dual role in both updating internal representations and internal models (Friston, 2005; Keller and Sterzer, 2024; Rao and Ballard, 1999).

Predictions about sensory input can be made based on a variety of sources of other information the brain has available. Certain stimuli can be predicted from the preceding sensory input – experimentally this can be used, e.g., in oddball paradigms to trigger stimulus history based prediction errors (Garrido et al., 2009). Predictions can be crossmodal. Certain sounds are associated with specific types of visual input. Experimentally visual stimuli can be coupled to sounds to then trigger audiovisual prediction errors in mouse visual cortex (Garner and Keller, 2022). Predictions can also be based on memory. Not seeing a stimulus in a specific spatial location can result in visuospatial prediction errors in mouse cortex (Fiser et al., 2016). However, the strongest predictor - by consistency of coupling - is self-generated movement. Every movement is directly coupled to sensory feedback throughout life.

An essential ingredient to the computation of visuomotor prediction error responses are motor-related predictions of visual feedback (Keller and Mrsic-Flogel, 2018). In the mouse, evidence for such predictions have come from the discovery of a strong motor-related modulation of activity in visual cortex (Keller et al., 2012; Niell and Stryker, 2010; Saleem et al., 2013). These motor-related signals are likely in part driven by motor-related predictions arriving from motor areas of cortex (Leinweber et al., 2017). However, the overall level of this motor-related activity is much higher than one would expect simply from predictions of visual feedback that are compared against visual input.

The more precise the prediction and comparison, the less motor-related activity should be detectable in visual cortex. This is evident in the auditory system of a songbird, which relies on very precise sensorimotor error detection for vocal learning, where there is much less motor modulation of auditory responses (Keller and Hahnloser, 2009). Similarly, there is very little movement related modulation of activity in visual cortex of non-human primates (Liska et al., 2024; Talluri et al., 2023). In humans, movement-related modulation of EEG activity in visual cortex has been recently described (Cheng and Nordin, 2025), but it is still unclear if visual responses are modulated by these signals. If indeed the precision of visuomotor coupling determines the amount of motor modulation of visual responses, we would expect to find less of it in humans than in the mouse.

Imprecise comparisons of motor related predictions and visual input should manifest as increases in motor-related activity and smaller prediction error response. Based on this, we would expect to find strong visuomotor mismatch responses in humans, similar to – or stronger than - those observed in mice (Keller et al., 2012). Based on this and prior work demonstrating that there are selective responses in human visual cortex to a mismatch between actual hand position and that of a virtual hand (Stanley and Miall, 2007), we expected to find strong EEG responses to a break in coupling between locomotion and visual feedback.

Results

Visual responses

To quantify visual responses in freely moving human participants we combined a virtual reality headset with an 8-electrode wireless EEG recording system (Methods; Figure 1A). Participants were shown a virtual environment that consisted of a large empty floor space with a square checkerboard pattern fixed to their viewing angle directly in front of them (Figure 1B). The checkerboard covered a visual angle of 53° and contained a red fixation dot in the center. Participants were asked to fixate on the red dot throughout all measurements. The checkerboard reversed white-black at intervals sampled from a uniform distribution between 2 and 4 s (Figure 1C). Recordings were split into a passive session during which participants were seated on a chair, and an active session during which they were instructed to walk around a rectangular floor area of 5 by 7 m. During the entire experiment we tracked the location of participants (Figure 1D).

EEG responses to visual stimulation are modulated by walking.

(A) Participant wearing the setup for wireless EEG recordings and a virtual reality headset. An Open BCI EEG electrode cap is combined with META QUEST 3. Inset: first person view of the virtual environment. Red fixation dot was enlarged in figure to make it visible. (B) A third person view of the virtual scene used to study visual responses to a reversing checkerboard pattern. The black line on the ground indicates the safety boundary within which the participant was instructed to walk around during the movement block of the paradigm. The grid indicates 1 by 1 m squares. (C) Participants viewed a reversing checkerboard in two conditions: during half of the trials, they remained seated, and during the other half they walked within a defined safety boundary. Checkerboard reversed colors at random intervals every 2 to 4 s. (D) 3D trajectory of an example participant during the visual experiment. Movement was recorded using accelerometers integrated into the VR headset. (E) Visually evoked potentials measured in the sitting condition on the occipital electrodes O1 and O2. Responses recorded on other electrodes are presented in Figure S2. Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red line indicates the time of the checkerboard reversal. P1 - positive peak at around 88 ms. (F) As in E, but while participants were walking. Inset: comparison of visual evoked potentials measured in the sitting E and walking F conditions. The horizontal bar above the plot marks time bins in which responses differ significantly (black: p < 0.05) or do not differ (gray: p > 0.05).

Walking triggered movement related artifacts in the EEG signals (Figure S1). The strength of these movement artifacts varied as a function of hairstyle and gait of the participants. All participants were encouraged to go ‘gentle’ to reduce this problem. We excluded all data with high movement related artifacts from further analysis (Methods; visual paradigm: 19 of 48 recordings, visuomotor mismatch paradigm: 15 of 48 recordings).

We could detect multiphasic responses to the checkerboard inversion in occipital EEG electrodes (Figures 1E and S2). These responses were comparable to previous reports (Drislane, 2007) and exhibited a positive peak at around 88 ms (P1). Interestingly, when participants were walking, the same responses showed an early negative deflection with a peak latency of 48 ms that preceded P1 (Figure 1F).

While participants were instructed to fixate on the red dot, it is possible that there are systematic differences in eye movements between walking and sitting conditions. Our setup did not allow for concurrent eye tracking; thus we cannot exclude the possibility that differences in eye movements contribute to the differences in EEG responses we observe. However, given the relatively rapid onset of response differences (40 ms), we suspect that stimulus triggered eye movement cannot account for these early differences.

Visuomotor mismatch responses

To measure visuomotor mismatch responses, participants were instructed to walk around in a virtual corridor (Figure 2A). The corridor was 1 m wide, 2.4 m high and of an oval shape covering approximately 7 by 5 m (Figure 2B). Virtual movement in the corridor was coupled to the movement of participants. We refer to this as closed loop. To trigger visuomotor mismatches, we briefly (0.5 s) halted the coupling between movement of the participants and visual feedback in the virtual corridor at random times (every 10 to 15 s; Methods; Figure 2C). Throughout all experiments, we tracked the virtual position of participants as they were walking around the corridor (Figure 2D).

Visuomotor mismatch elicits strong EEG responses.

(A) Participant navigating a virtual tunnel. Inset: first person view of the tunnel. (B) Third person view of the virtual environment used to study visuomotor mismatch responses. The grid indicates 1 by 1 m squares. (C) In closed loop sessions, the walking speed of participants was coupled to movement in the virtual corridor. Visuomotor mismatches were introduced by briefly halting the visual flow for 0.5 s. Following each visuomotor mismatch event, the view was updated to the participants’ current position (brief peaks in visual flow speed after the mismatch event) and visuomotor coupling was resumed. (D) Trajectory of an example participant during the visuomotor mismatch paradigm. (E) Responses to visuomotor mismatches recorded from occipital electrodes (O1 and O2). Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset and offset of the visuomotor mismatch. The gray shaded areas mark the analysis windows used to quantify response strength in G and H. Analysis windows of 100 ms were centered on the peak of the visuomotor mismatch response and the visual flow re-onset response. (F) As in E, but for visual flow playback halt responses recorded from occipital electrodes. (G) Comparison of the response strength to visuomotor mismatches and playback halts. Here and elsewhere: boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from one participant. ***: p<0.001. One data point not shown (Mismatch response, at −9.5 μV). See Table S1 for all statistical information. (H) Comparison of the response strength to visual flow re-onset events following visuomotor mismatches in the closed loop condition and playback halts in the open loop condition. n.s.: not significant. See Table S1 for all statistical information. (I) Average walking speed of participants during visuomotor mismatches. Inset: Temporally expanded view. The horizontal bar above the plot marks time bins where walking speed differs significantly (black: p < 0.05) or does not differ (gray: p > 0.05) from baseline.

Anecdotally, the perception some participants reported experiencing during these visuomotor mismatches was a very salient sudden movement of the entire visual world. In the words of one participant: “The world suddenly flew forward! – Are you printing this? – Hi Mom!”. This was reminiscent of a case report of a patient with a lesion to the lateral rectus muscle described by von Helmholtz (von Helmholtz, 1867). The lateral rectus muscle moves the eye laterally. The patient, upon attempting to move the affected eye laterally, reported seeing the world rapidly moving in the direction of the intended eye movement. One interpretation of this is that the combination of movement and absence of resulting visual flow change results in the perception that the world must be moving.

In the EEG recordings of occipital electrodes, we found strong responses triggered by these visuomotor mismatches (Figure 2E). Responses were dominated by a positive component peaking at 180 ms. Following the 0.5 s halting of the visual stimulus, the virtual location was updated to match the actual location of participants, and the visual flow coupling was resumed. This resulted in a sudden change of visual stimulus combined with a visual flow onset and drove what is likely a visual response in the EEG signal (Figure 2E).

To quantify how much of the visuomotor mismatch response can be explained by visual input alone, we exposed participants to a replay of the visual flow that was self-generated in the preceding closed loop session, including the visual flow halts that constitute visuomotor mismatch in the closed loop condition. We refer to this as open loop. For these experiments, participants were seated on a chair. To reduce the likelihood of triggering nausea in participants, roll and pitch movements of the head were removed from this replay. We found measurable responses to visual flow playback halts (Figure 2F), but these were much smaller than visuomotor mismatch responses (Figure 2G). In the few participants who volunteered to experience playback with pitch and roll movements, playback halt responses were not different from those in the reduced playback condition (Figure S3). The difference in response amplitude between mismatch and playback response could not be explained by a simple movement related gain of visual responses, as the response to the re-onset of visual flow was similar in both closed and open loop conditions (Figure 2H). Note that the mismatch responses and playback halt responses shown here do not come from fully overlapping pools of participants (Table S1). A subset of sessions had to be excluded because of recording noise (visuomotor mismatch: 15 of 48, playback: 4 of 48), while some participants aborted the playback recording due to developing nausea (11 of 48; Methods). A participant-wise comparison of mismatch and playback halt responses is shown in Figure S4.

Although participants reduced their walking speed in response to the visuomotor mismatch, this reduction was merely a trend and did not reach statistical significance (Figure 2I). However, given the comparably slow time course of reduction in walking speed, these behavioral changes cannot explain visuomotor mismatch responses.

Distribution of visuomotor mismatch response strength

In mice, visuomotor mismatch responses originate in primary visual cortex and propagate to a network of areas across dorsal cortex (Heindorf and Keller, 2023; Takeuchi et al., 2024). Thus, we expected to find larger and faster responses in the two occipital electrodes. To test this, we compared the visuomotor mismatch responses across the eight recorded locations in both response strength and timing (Figures 3A and 3B). We indeed found that the strongest responses were observed at occipital electrodes O1 and O2, though significant responses were also present at frontal (Fp1) and central (C3) sites (Figure 3C). However, while there was a trend toward faster responses in occipital cortex, we found no evidence of differences in response latency across the different electrodes (Figures 3D and 3E).

Visuomotor mismatch responses are most prominent in occipital cortex.

(A) Top down view of EEG electrode locations on the head. (B) Visuomotor mismatch responses measured on electrodes shown in A. Solid black lines represent the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (C) Comparison of the response strength to visuomotor mismatches measured on electrodes shown in A. Average response strength was calculated within a 100 ms window centered on the peak of the average visuomotor mismatch response across all electrodes. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from individual participant. ***: p<0.001, **: p<0.01, *: p<0.05, n.s.: not significant. See Table S1 for all statistical information. (D) The responses shown in B, averaged over pairs of electrodes and overlaid. Solid black lines represent mean and shading SEM across electrodes. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (E) Comparison of the latency to half maximum response for the 4 electrode pairs. n.s.: not significant. See Table S1 for all statistical information.

Comparison of visual responses and visuomotor mismatch responses

The time course of visual responses (Figure 1E) and visuomotor mismatch responses (Figure 2E) differed in that they appeared to exhibit reversed polarity (Figure 4A). Despite this difference in polarity, the initial peaks occurred at similar latencies. The total power of the visuomotor response was higher than that of the visual response (Figure 4B) but given that we did not optimize either the visuomotor mismatch stimulus nor the visual stimulus for maximum response, this comparison is more qualitative than quantitative in nature. We include it here primarily to illustrate how the visuomotor mismatch response compares to a more conventional visual response.

Visuomotor mismatch responses have reversed polarity and more power compared to visual responses.

(A) Comparison of visual (Figure 1E) and visuomotor mismatch (Figure 2E) responses recorded from occipital electrodes. Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset (visual and mismatch) and offset (mismatch) of the stimuli. (B) Comparison of the power of visual and visuomotor mismatch responses, calculated within a 0 - 0.5 s time window following stimulus onset. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from individual participant. ***: p<0.001. See Table S1 for all statistical information.

Comparison of visuomotor mismatch responses and auditory oddball mismatch responses

To further contrast the strength of visuomotor mismatch to other frequently used prediction error signals, we also recorded EEG responses in an auditory oddball paradigm. Mismatch responses recorded with this paradigm can be thought of as a stimulus history prediction error (Garrido et al., 2009; Näätänen et al., 2007). We implemented an oddball paradigm composed of a sequence of identical tones that was occasionally disrupted by a tone of a different frequency (Figure 5A). We used pure tones of 1 kHz and 1.2 kHz frequency, and alternated in blocks which tone was standard and which was the deviant (Methods). To enable direct comparison with visuomotor mismatch responses, recordings were made from occipital electrodes O1 and O2. Participants watched short silent movies in the VR headset while listening to the tone sequences delivered in headphones. To isolate the oddball driven mismatch response, we subtracted the response to the tone when presented as a standard from the response to the same tone when presented as an deviant. The resulting difference waveform closely resembled those reported in previous studies (Figures 5B and 5C) (Näätänen et al., 2007). Comparing these oddball mismatch responses to the visuomotor mismatch response, we found that both had similar dynamics with a dominant positive peak at around 180 ms (Figure 5D). Comparing the total power of the two responses, we found that visuomotor mismatch responses were significantly larger than oddball mismatch responses (Figure 5E). Also here the comparison of total power is more qualitative than quantitative as neither stimulus was optimized for maximum EEG response, and we include it to illustrate how typical variants of these responses compare.

Visuomotor mismatch responses are larger than auditory oddball mismatch responses but have similar temporal dynamics

(A) Design of the auditory oddball paradigm with examples of the silent films participants were exposed to. (B) Top: Auditory responses to the 1 kHz tone presented as a standard versus as a deviant. Bottom: Oddball mismatch response calculated by subtracting the average response over trials in which the tone was presented as a standard from those when the tone was presented as a deviant. Solid black lines represent the mean, and shading indicates the SEM across participants. Dashed vertical red line is the onset of the auditory stimulus. (C) As in B, but for the 1.2 kHz tone. (D) Comparison of visuomotor mismatch, oddball mismatch (average over data shown in panel B and C) and playback halt responses recorded from occipital electrodes. Solid black lines represent the mean response, with shading indicating SEM across electrodes. Dashed vertical red lines are onset (visual, mismatch) and offset (mismatch) of the stimuli. (E) Comparison of the power of visuomotor mismatch, oddball mismatch response (average over data shown in panel B and C) and playback halt responses, calculated within a 0 s - 0.5 s time window following stimulus onset. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from individual participant. ***: p<0.001; *p<0.05. See Table S1 for all statistical information.

Discussion

The utility of paradigms for measuring prediction errors in humans primarily comes from their relevance to psychiatric disorders, where disruptions in prediction error processing are widely implicated (Adams et al., 2013; Fletcher and Frith, 2009; Kirihara et al., 2020; Qela et al., 2025; Sterzer et al., 2018). Various types of prediction error responses are being explored as potential biomarkers in clinical research, serving to monitor disease progression, evaluate symptoms, and provide insight into the underlying neural mechanisms of these conditions.

The most prominent of these prediction errors is mismatch negativity (MMN). MMN is the first component of the oddball mismatch response (Näätänen et al., 1978). MMN has been shown to be reduced or abnormal in patients with schizophrenia spectrum disorders (Todd et al., 2012; Umbricht and Krljes, 2005), autism spectrum disorders (Chen et al., 2020; Dunn et al., 2008; Schwartz et al., 2018), speech disorders (El Hatal de Souza et al., 2020), depression (Tseng et al., 2021), and bipolar disorders (Chitty et al., 2013). More generally, it is thought that many of these disorders can be related to changes in predictive processing. In the case of schizophrenia, for example, the symptoms are thought to arise from an imbalance in the strength of high and low level predictions (priors) (Schmack et al., 2017, 2015a, 2015b, 2013). Further testing these ideas will require experiments that trigger identified functional responses in humans, for which we have a circuit level understanding based on cell type specific recordings.

We were able to measure robust visuomotor mismatch responses in freely moving humans. These signals cannot be attributed to changes in participants’ behavior or to simple visual offset responses. Notably, the visuomotor mismatch responses exhibited a markedly different temporal profile compared to purely visual responses. What might account for this difference? One possibility is that visuomotor mismatch signals may rely more strongly on input from other cortical areas than visual responses and are thus delayed relative to visual responses. A more interesting interpretation is that visual responses and visuomotor mismatch responses are both prediction errors of different types. A central tenet of the cortical circuit for predictive processing is the split into separate populations of neurons that compute positive and negative prediction errors (Keller and Mrsic-Flogel, 2018; Rao and Ballard, 1999). In this interpretation, a visuomotor mismatch response is a negative prediction error, while the response to a visual stimulus is a positive prediction error.

Visuomotor prediction errors are likely computed in layer 2/3 of mouse visual cortex (Jordan and Keller, 2020), and are also preferentially detectable in superficial layers of human visual cortex (Thomas et al., 2024). In the mouse, positive and negative prediction errors are likely computed in separate populations of neurons in visual cortex (Keller and Mrsic-Flogel, 2018; O’Toole et al., 2023). These two cell types have a different depth distribution, with negative prediction error neurons more superficial and positive prediction error neurons located deeper in cortex (O’Toole et al., 2023). Given that EEG signals are thought to exhibit positive or negative deflections as a function of depth of the source (Cohen, 2017; Kirschstein and Köhling, 2009), the polarity reversal of visual and visuomotor mismatch responses (Figure 4) may be the consequence of different populations of cells being activated and inhibited. Visuomotor mismatch should activate negative prediction error neurons and inhibit positive prediction error neurons, while the reverse is the case for a visual stimulus.

The fact that there is also a strong visual response to the visual flow re-onset following visuomotor mismatch means that our visuomotor mismatch paradigm might allow us to measure both negative and positive prediction error responses. More intriguingly, there is a differential modulation by walking of the two responses (Figures 2G and 2H). In theory, visuomotor mismatch drives a combination of two negative prediction errors. One is based on a movement related prediction. The other on a stimulus-history related prediction. Both the self-motion as well as the ongoing visual flow would predict a continuation of visual flow. A visuomotor mismatch violates both predictions. During passive observation there is no movement-related prediction, but the stimulus history-based prediction is still violated.

Intriguing is the similarity in timing and polarity of the playback halt and the difference between oddball and standard response (Figure 5D). An oddball response is always a combination of both a positive and a negative prediction error. There is a negative prediction error for the absence of the expected standard tone as well as a positive prediction error for the presence of the unexpected oddball tone. Thus, it is conceivable that the subtraction of standard response from the oddball response, isolates the component of the oddball response that corresponds to a negative prediction error.

One of the key challenges in systems neuroscience is translating findings from animal models to humans. Although recent animal studies have provided detailed insights into the circuit level implementation of predictive processing in the cortex (Keller and Mrsic-Flogel, 2018), and psychological and psychiatric conditions have long been described within this framework (Keller and Sterzer, 2024; Sterzer et al., 2018), translation to human research has been limited by a lack of methods to record cell type specific signals in human experiments. With an understanding from animal models of how positive and negative prediction errors are computed, and which cortical layers are involved, an approach based on functionally identified responses might be most promising. One step in this direction is to use experimental paradigms that can cleanly separate e.g. positive and negative prediction error responses. As argued above, we speculate that the visuomotor mismatch paradigm as we have used it here, is one possible way to achieve this.

Methods

Participants

The study was approved by Ethikkommission Nordwest- und Zentralschweiz (Project-ID: 2024-02458). Participants signed a written informed consent before participation and were not financially compensated for their participation. 46 healthy adults (19 males, 27 females) participated in the study, ranging in age from 18 to 65 years. The age distribution was as follows: 19 participants aged 18-30, 18 aged 31-40, 5 aged 41-50, and 4 aged 51-65. Two participants went through the recordings twice, resulting in 48 recorded sessions in total. None of the participants reported a prior diagnosis of movement disorders, vestibular dysfunction, or epilepsy. Experience with virtual reality (VR) technology ranged from beginner to advanced, with most participants reporting minimal prior experience.

EEG recordings in humans

We integrated a wireless EEG recording system with a VR headset. The EEG recording system was composed of a wet electrode cap and a Cyton biosensing board from OpenBCI. This allowed us to record 8 EEG channels (45 sessions: FP1, FP2, C3, C4, P3, P4, O1 and O2; 3 sessions: FP1, FP2, T3, T4, T5, T6, O1 and O2) at a 250 Hz sampling rate. Electrode labels follow the international 10-20 system: FP = frontopolar, C = central, P = parietal, O = occipital, and T = temporal; odd numbers indicate the left hemisphere, and even numbers the right (Figure 3A). All EEG data were wirelessly (via Bluetooth) transmitted to a nearby computer. The VR headset was a Meta Quest 3, with a virtual environment developed in the Unity engine (Unity Technologies). Virtual 3D objects were designed in Fusion 360 (Autodesk). Synchronization between EEG recording and VR headset was performed by connecting an auditory output of the VR headset to the Cyton board to exchange synchronization triggers. This allowed us to synchronize EEG data with VR events offline. Auxiliary signals, including the participant’s position, stimulus trigger timing, and type, were recorded directly on the headset at a sampling rate of approximately 100 Hz.

Visual responses

Visual stimuli were presented using the VR headset. We used a reversing square checkerboard stimulus to drive visually evoked potentials. The checkerboard reversed colors at random intervals (between 2 and 4 s). In virtual space, the checkerboard measured 0.5 by 0.5 m and was positioned 0.5 m in front of participants’ eyes. This resulted in a horizontal and vertical coverage by the checkerboard of approximately 53° of visual angle. Each visual stimulation session lasted 4 min, during which the checkerboard reversed colors between 78 and 82 times. For half of the session, participants viewed the stimulus while seated; for the other half, they were instructed to walk freely within a 7 by 5 m empty floor space. The order of sitting and walking conditions was randomized across participants.

Visuomotor mismatch responses

For the experiments measuring visuomotor mismatch responses, we used a 3D virtual corridor with vertical gratings on the walls. The corridor measured 1 m in width, 2.4 m in height, and had an oval shape of 7 by 5 m. Visuomotor mismatches were introduced at random intervals every 10 to 15 s as participants walked through the corridor. The session lasted 5 min and included between 22 and 26 visuomotor mismatch events. During these events, the coupling between the participant’s movement and the visual feedback in the VR headset was briefly interrupted, i.e. the visual scene was frozen for 0.5 s. Because participants continued to move during this time, they were teleported to their current position in the virtual space following visuomotor mismatch event.

Playback halt responses

To quantify how much of the visuomotor mismatch response could be explained by visual input alone, participants were asked to passively observe a replay of the visual flow they had self-generated during the preceding closed loop session. We quickly learned, however, that watching 5 minutes of playback in the VR headset triggered nausea in most participants. Thus, we started experimenting with changes to the playback to minimize the risk of triggering nausea. One such modification we settled on to use for experiments was a playback version that omitted pitch and roll movements of the head. Full playback involved six degrees of freedom (6DOF): 3D position in space, plus pitch, yaw, and roll angles of the head. The constrained playback included only four degrees of freedom (4DOF): the 3D position in space, but only yaw movements of the head. A subset of participants volunteered to view the full playback (Figure S3). Playback sessions were conducted while participants were seated and lasted 5 min, matching the duration of the closed loop session. Approximately 23% of participants reported strong nausea during the beginning of the 4DOF playback session, at which point the recordings were terminated.

Mismatch negativity

To measure mismatch negativity we used an auditory oddball paradigm. Auditory stimuli consisted of two pure tones: 1 kHz and 1.2 kHz. They were presented at 60 to 70 dB sound pressure level in a counterbalanced design, in which each tone served as the standard in one condition and as the deviant in the other. Tones were 50 ms in duration, including 5 ms linear rise and fall ramps. Stimuli were delivered in a pseudorandom order, with deviant tones comprising 15% of all trials and preceded by at least five standard tones. The inter-trial interval was selected from a uniform distribution of between 500 and 600 ms. Participants were seated and listened to the tones while watching a silent movie that was not related to the tone sequences.

EEG Signal Analysis

Data analysis was done using custom-written MATLAB scripts. EEG signals were band-pass filtered between 0.2 and 100 Hz. To remove power line noise, a band-stop filter was applied between 49 and 51 Hz. Movement of the participants triggered all varieties of movement related artifacts in the EEG recordings. The strength of these artifacts depended on a variety of factors: impedance of the electrodes, hair style of participants, gait pattern, and likely others. To reduce data contaminated by excessive movement artifacts, trials with a maximum absolute response amplitude exceeding 100 μV were discarded from further analysis (Figure S5). Data from each electrode were included in the final analysis only if at least 15 triggers remained after exclusion of triggers with excessive movement artifacts (Table S1). To compare average response strength across conditions, a 100 ms analysis window was used, centered on the peak of the respective responses: the visuomotor mismatch event recorded at occipital electrodes O1 and O2 (Figures 2E, 2F, and S4), the mean visuomotor mismatch response across all electrodes (Figure 3C), the mean visual response across all electrodes (Figure S2C) or playback halt response in the 6DOF condition (Figure S3B). Signal power was compared by calculating the mean squared amplitude within a 0 - 0.5 s analysis window following stimulus onset (Figures 4B and 5E).

Statistical tests

All statistical analyses were conducted using hierarchical bootstrap (Saravanan et al., 2020). Bootstrap resampling enables statistical comparisons across conditions without assuming a specific distribution of the EEG data. For analysis in Figures 1, 2, 4, 5, S3, and S4, we averaged signals from electrodes O1 and O2, and treated the result as a single data point per participant. For the analysis shown in Figures 3C, 3E, and S2C we included signals from all electrodes and used a nested bootstrap to account for multiple data points originating from the same participant. We first resampled the data with replacement at the level of participants, followed by resampling at the level of electrodes. For each resampled population, we computed the mean response and repeated this procedure 10000 times. The p value was estimated as the fraction of bootstrap samples in which the sample mean violated the tested hypothesis. See Table S1 for all information on number of participants or electrodes used for all analyses shown.

eLife Assessment

This valuable study demonstrates that self-motion strongly affects neural responses to visual stimuli, comparing humans moving through a virtual environment to passive viewing. However, evidence that the modulation is due to prediction is incomplete as it stands, since participants may come to expect visual freezes over the course of the experiment. This study bridges human and rodent studies on the role of prediction in sensory processing, and is therefore expected to be of interest to a large community of neuroscientists.

Reviewer #1 (Public review):

In this paper, the authors wished to determine human visuomotor mismatch responses in EEG in a VR setting. Participants were required to walk around a virtual corridor, where a mismatch was created by halting the display for 0.5s. This occurred every 10-15 seconds. They observe an occipital mismatch signal at 180 ms. They determine the specificity of this signal to visuomotor mismatch by subsequently playing back the same recording passively. They also show qualitatively that the mismatch response is larger than one generated in a standard auditory oddball paradigm. They conclude that humans therefore exhibit visuomotor mismatch responses like mice, and that this may provide an especially powerful paradigm for studying prediction error more generally.

Asking about the role of visuomotor prediction in sensory processing is of fundamental importance to understanding perception and action control, but I wasn't entirely sure what to conclude from the present paradigm or findings. Visuomotor prediction did not appear to have been functionally isolated. I hope the comments below are helpful.

(1) First, isolating visuomotor prediction by contrasting against a condition where the same video stream is played back subsequently does not seem to isolate visuomotor prediction. This condition always comes second, and therefore, predictability (rather than specifically visuomotor predictability) differs. Participants can learn to expect these screen freezes every 10-15 s, even precisely where they are in the session, and this will reduce the prediction error across time. Therefore, the smaller response in the passive condition may be partly explained by such learning. It's impossible to fully remove this confound, because the authors currently play back the visual specifics from the visuomotor condition, but given that the visuomotor correspondences are otherwise pretty stable, they could have an additional control condition where someone else's visual trace is played back instead of their own, and order counterbalanced. Learning that the freezes occur every 10-15 s, or even precisely where they occur, therefore, could not explain condition differences. At a minimum, it would be nice to see the traces for the first and second half of each session to see the extent to which the mismatch response gets smaller. This won't control for learning about the specific separations of the freezes, but it's a step up from the current information.

(2) Second, the authors admirably modified their visual-only condition to remove nausea from 6 df of movement (3D position, pitch, yaw, and roll). However, despite the fact it's far from ideal to have nauseous participants, it would appear from the figures that these modifications may have changed the responses (despite some pairwise lack of significance with small N). Specifically, the trace in S3 (6DOF) and 2E look similar - i.e., comparing the visuomotor condition to the visual condition that matches. Mismatch at 4/5 microvolts in both. Do these significantly differ from each other?

(3) It generally seems that if the authors wish to suggest that this paradigm can be used to study prediction error responses, they need to have controlled for the actions performed and the visual events. This logic is outlined in Press, Thomas, and Yon (2023), Neurosci Biobehav Rev, and Press, Kok, and Yon (2020) Trends Cogn Sci ('learning to perceive and perceiving to learn'). For example, always requiring Ps to walk and always concurrently playing similar visual events, but modifying the extent to which the visual events can be anticipated based on action. Otherwise, it seems more accurately described as a paradigm to study the influence of action on perception, which will be generated by a number of intertwined underlying mechanisms.

More minor points:

(1) I was also wondering whether the authors may consider the findings in frontal electrodes more closely. Within the statistical tests of the frontal electrodes against 0, as displayed in Figure 3c, the insignificance of the effect of Fp2 seems attributable to the small included sample size of just 13 participants for this electrode, as listed in Table S1, in combination with a single outlier skewing the result. The small sample size stands out especially in comparison to the sample size at occipital electrodes, which is double and therefore enjoys far more statistical power. It looks like the selected time window is not perfectly aligned for determining a frontal effect, and also the distribution in 3B looks like responses are absent in more central electrodes but present in occipital and frontal ones. I realise the focus of analysis is on visual processing, but there are likely to be researchers who find the frontal effect just as interesting.

(2) It is claimed throughout the manuscript that the 'strongest predictor (of sensory input) - by consistency of coupling - is self-generated movement'. This claim is going to be hard to validate, and I wonder whether it might be received better by the community to be framed as an especially strong predictor rather than necessarily the strongest. If I hear an ambulance siren, this is an especially strong predictor of subsequent visual events. If I see a traffic light turn red, then yellow, I can be pretty certain what will happen next. Etc.

(3) The checkerboard inversion response at 48 ms is incredibly rapid. Can the authors comment more on what may drive this exceptionally fast response? It was my understanding that responses in this time window can only be isolated with human EEG by presenting spatially polarized events (cf. c1, e.g., Alilovic, Timmermans, Reteig, van Gaal, Slagter, 2019, Cerebral Cortex)

Reviewer #2 (Public review):

Summary:

This study investigates whether visuomotor mismatch responses can be detected in humans. By adapting paradigms from rodent studies, the authors report EEG evidence of mismatch responses during visuomotor conditions and compare them to visual-only stimulation and mismatch responses in other modalities.

Strengths:

(1) The authors use a creative experimental design to elicit visuomotor mismatch responses in humans.

(2) The study provides an initial dataset and analytical framework that could support future research on human visuomotor prediction errors.

Weaknesses:

(1) Methodological issues (e.g., volume conduction, channel selection, lack of control for eye movements) make it difficult to confidently attribute the observed mismatch responses to activity in visual cortical regions.

(2) A very large portion of the data was excluded due to motion artefacts, raising concerns about statistical power and representativeness. The criteria for trial inclusion and the number of accepted trials per participant appear arbitrary and not justified with reference to EEG reliability standards.

(3) The comparison across sensory modalities (e.g., auditory vs. visual mismatch responses) is conceptually interesting, but due to the choice of analyzing auditory mismatch responses over occipital channels, it has limited interpretability.

The authors successfully demonstrate that visuomotor mismatch paradigms can, in principle, be applied in human EEG. However, due to the issues outlined above, the current findings are relatively preliminary. If validated with improved methodology, this approach could significantly advance our understanding of predictive processing in the human visual system and provide a translational bridge between rodent and human work.

Reviewer #3 (Public review):

Summary:

Solyga, Zelechowski, and Keller present a concise report of an innovative study demonstrating clear visuomotor mismatch responses in ambulating humans, using a mobile EEG setup and virtual reality. Human subjects walked around a virtual corridor while EEGs were recorded. Occasionally, motion and visual flow were uncoupled, and this evoked a mismatch response that was strongest in occipitally placed electrodes and had a considerable signal-to-noise ratio. It was robust across participants and could not be explained by the visual stimulus alone.

Strengths:

This is an important extension of their prior work in mice, and represents an elegant translation of those previous findings to humans, where future work can inform theories of e.g., psychiatric diseases that are believed to involve disordered predictive processing. For the most part, the authors are appropriately circumspect in their interpretations and discussions of the implications. I found the discussion of the polarity differences they found in light of separate positive and negative prediction errors, intriguing.

Weaknesses:

The primary weaknesses rest in how the results are sold and interpreted.

Most notably, the interpretation of the results of the comparison of visuomotor mismatches to the passive auditory oddball induced mismatch responses is inappropriate, as suboptimal electrode choices, unclear matching of trial numbers, and other factors. To clarify, regarding the auditory oddball portion in Figure 5, the data quality is a concern for the auditory ERPs, and the choice of Occipital electrodes is a likely culprit. Typically, auditory evoked responses are maximal at Cz or FCz, although these contacts don't seem to be available with this setup. In general, caution is warranted in comparing ERP peaks between two different sensory modalities - especially if attention is directed elsewhere (to a silent movie) during one recording and not during the other. The authors discuss this as a purely "qualitative" comparison in the text, which is appreciated, and do acknowledge the limitations within the results section, but the figure title and, importantly, the abstract set a different tone. At least, for comparisons between auditory mismatch and visuomotor mismatch, trial numbers need to be equated, as ERP magnitude can be augmented by noise (which reduces with increased numbers of trials in the average). And more generally, the size of the mismatch event at the scalp does not scale one-to-one with the size at the level of the neural tissue. One can imagine a number of variables that impact scalp level magnitudes, which are orthogonal to actual cortex-level activation - the size, spread, and polarity variance of the activated source (which all would diminish amplitude at the scalp due to polyphasic summation/cancelation). The variance of phase to a stimulus across trials (cross trial phase locking) vs magnitude of underlying power - the former, in theory, relates to bottom-up activity and the latter can reflect feedback (which has more variability in time across trials; the distance of the scalp electrode from the activated tissue (which, for the auditory system, would be larger (FCz to superior temporal gyrus) than for the visual system (O1 to V1/2)). None of this precludes the inclusion of the auditory mismatch, which is a strength of the study, but interpretations about this supporting a supremacy of sensory-motor mismatch - regardless of validity - are not warranted. I would recommend changing the way this is presented in the abstract.

Otherwise, the data are of adequate quality to derive most of their conclusions.

The authors claim that the mismatch responses emanate from within the occipital cortex, but I would require denser scalp coverage or a demonstration of consistent impedances across electrodes and across subjects to make conclusions about the underlying cortical sources (especially given the latencies of their peaks). In EEG, the distribution of voltage on the scalp is, of course, related to but not directly reflective of the distribution of the underlying sources. The authors are mostly careful in their discussion of this, but I would strongly recommend changing the work choice of "in occipital cortex" to "over occipital cortex" or even "posteriorly distributed". Even with very dense electrode coverage and co-registration to MRIs for the generation of forward models that constrain solutions, source localization of EEG signals is very challenging and not a simple problem. Given the convoluted and interior nature of human V1, the ability to reliably detect early evoked responses (which show the mismatch in mouse models) at the scalp in ERP peaks is challenging - especially if one is collapsing ERPs across subjects. And - given the latency of the mismatch responses, I'd imagine that many distributed cortical regions contribute to the responses seen at the scalp.

I think that Figure 3C, but as a difference of visual mismatch vs halting flow alone (in the open loop) might be additionally informative, as it clarifies exactly where the pure "mismatch" or prediction error is represented.

As a suggestion, the authors are encouraged to analyse time-frequency power and phase locking for these mismatch responses, as is common in much of the literature (see Roach et al 2008, Schizophrenia Bulletin). This is not to say that doing so will yield insights into oscillations per se, but converting the data to the time-frequency domain provides another perspective that has some advantages. It fosters translations to rodent models, as ERP peaks do not map well between species, but e.g., delta-theta power does (see Lee et al 2018, Neuropsychopharmacology; Javitt et al 2018, Schizophrenia research; Gallimore et al 2023, Cereb Ctx). Further, ERP peaks can be influenced by the actual neuroanatomy of an individual (especially for quantifying V1 responses). Time frequency analyses may aid in interpreting the "early negative deflection with a peak latency of 48 ms " finding as well.

Finally, the sentence in the abstract that this paradigm " can trigger strong prediction error responses and consequently requires shorter recording 20 times would simplify experiments in a clinical setting" is a nice setup to the paper, but the very fact that one third of recordings had to be removed due to movement artifact, and that hairstyle modulates the recording SnR, is reason that this paradigm, using the reported equipment, may have limited clinical utility in its current form. Further, auditory oddball paradigms are of great clinical utility because they do not require explicit attention and can be recorded very quickly with no behavioral involvement of a hospitalized patient. This should be discussed, although it does not detract from the overall scientific importance of the study. The authors should reconsider putting this statement in the abstract.

Author response:

The following is the authors’ response to the original reviews.

We thank you for the time you took to review our work and for your feedback! The main changes to the manuscript are:

(1) We have performed additional experiments to increase the number of recordings from frontal and occipital electrodes (previously 51 (occipital: O1+O2) and 26 (frontal: Fp1+Fp2), now 133 and 102). The additional data have strengthened many of our results, including for example the trend for a latency difference between occipital and frontal electrodes that was likely underpowered and is now significant (Figure 3E). We have updated all relevant figures to include the additional data (Figures 2–6, Figure S4, Figure S5). None of the main conclusions have changed.

(2) As suggested by reviewer 1, we have conducted additional experiments to rule out the possibility that the observed effects were driven by the temporal order of open and closed loop sessions (new Figure S6). We also found another 9 participants who were willing to go on the ‘vomit comet’ of six degrees of freedom (6DOF) playback (previously 5, now 14). These data have further strengthened our conclusion that playback halt responses in 4DOF and 6DOF playback are not substantially different (Figure S4).

(3) To address the point of reviewers 2 and 3, that mismatch negativity (MMN) responses would be larger on temporal electrodes, we conducted additional experiments in which we also recorded from temporal electrodes T3–T6. We have now added a comparison of visuomotor mismatch and MMN responses on T3–T6 electrodes as Figures S8–S9. On all electrodes, visuomotor mismatch responses were larger than MMN responses.

(4) As suggested by reviewer 1, we have added an analysis of the experience-dependent changes in mismatch responses comparing frontal and occipital responses early and late in the session (new Figure 4).

(5) As suggested by reviewer 2, we conducted additional experiments in an independent cohort of participants (note, without concurrent EEG) to measure eye movements triggered by visuomotor mismatches. We found eye-movement speed and blink/eye-closure changes, but these had longer latency than visuomotor mismatch responses (Figure S7).

(6) Finally, as suggested by reviewers 2 and 3, we applied independent component (ICA) and time–frequency analyses to the EEG data. We show these results and explain why they are not applicable or useful in our case in the responses below.

Please note, during the revision, we found that a part of our analysis used a bandpass of 0.2-100 Hz while a 1-100 Hz bandpass filter was used elsewhere. This has now been standardized to a 1-100 Hz bandpass filter, and the corresponding methods were updated. This resulted in no relevant changes to the figures. Additionally, the 50 Hz band-stop filter was erroneously described in the methods as 49-51 Hz. The filter used was 40-60 Hz, and the methods have been updated to reflect this.

Reviewer #1 (Public review):

In this paper, the authors wished to determine human visuomotor mismatch responses in EEG in a VR setting. Participants were required to walk around a virtual corridor, where a mismatch was created by halting the display for 0.5s. This occurred every 10-15 seconds. They observe an occipital mismatch signal at 180 ms. They determine the specificity of this signal to visuomotor mismatch by subsequently playing back the same recording passively. They also show qualitatively that the mismatch response is larger than one generated in a standard auditory oddball paradigm. They conclude that humans therefore exhibit visuomotor mismatch responses like mice, and that this may provide an especially powerful paradigm for studying prediction error more generally.

Asking about the role of visuomotor prediction in sensory processing is of fundamental importance to understanding perception and action control, but I wasn't entirely sure what to conclude from the present paradigm or findings. Visuomotor prediction did not appear to have been functionally isolated. I hope the comments below are helpful.

(1) First, isolating visuomotor prediction by contrasting against a condition where the same video stream is played back subsequently does not seem to isolate visuomotor prediction. This condition always comes second, and therefore, predictability (rather than specifically visuomotor predictability) differs. Participants can learn to expect these screen freezes every 10-15 s, even precisely where they are in the session, and this will reduce the prediction error across time. Therefore, the smaller response in the passive condition may be partly explained by such learning. It's impossible to fully remove this confound, because the authors currently play back the visual specifics from the visuomotor condition, but given that the visuomotor correspondences are otherwise pretty stable, they could have an additional control condition where someone else's visual trace is played back instead of their own, and order counterbalanced. Learning that the freezes occur every 10-15 s, or even precisely where they occur, therefore, could not explain condition differences. At a minimum, it would be nice to see the traces for the first and second half of each session to see the extent to which the mismatch response gets smaller. This won't control for learning about the specific separations of the freezes, but it's a step up from the current information.

In theory, it is correct that the open loop (playback) session is predictable. However, this is relatively unrealistic. The open loop session is a 5-minute sequence that participants have only experienced once before, when they were generating it in the closed loop session a couple of minutes earlier. It is unlikely that participants would remember the entire sequence to a precision of less than a second, which is what they would need to predict the mismatch event. However, the reviewer is correct that it is possible that the mismatch events lose salience with time, for example as a consequence of participants losing interest in the task with time, or by undergoing some form of adaptation. To address this, we repeated the experiments with the sequence of closed and open loop sessions reversed (Figures S6A-S6C), and we analyzed the responses as a function of time within the session (Figures S6D and S6E), as suggested.

The reversed-order design consisted of (1) open loop session: a playback, in which participants viewed the recorded closed loop session of a previous participant. This was followed by (2) a closed loop session, in which participants actively walked through the tunnel and experienced visuomotor mismatch events. Using this design, we again found that responses in the closed loop session were significantly larger than in the open loop session (Figures S6A-S6C).

In addition, we analyzed both new and previously collected data as a function of time in the session. We computed moving average responses across 10 mismatch or playback halt trials at different percentages of progress through the paradigm (Figures S6D and S6E). This analysis revealed no consistent experience-dependent changes that could account for the observed differences between closed and open loop session. While there was indeed some form of experience dependent attenuation of visuomotor mismatch responses (see new Figure 4), the difference at the transition from mismatch to playback halt (and vice versa) far exceeded these adaptation effects (Figures S6D and S6E). This analysis was performed only on data from participants for whom we had both closed and open loop sessions and met our inclusion criteria.

We used a similar analysis to test whether early and late responses within a session systematically differed (new Figure 4). Here, to maximize the chance of finding a difference, we compared early (first five) and late (last five) trials. Behaviorally, participants reduced their walking speed following mismatch events, with a significantly larger reduction during early trials (14.3%) than during late trials (5.7%) (Figure 4A). Neural responses mirrored this pattern primarily on frontal electrodes: frontal activity showed a clear attenuation from early to late trials (Figure 4B), consistent with the reduction in behavioral responses. In contrast, changes on occipital electrodes were much smaller between early and late trials (Figure 4C-4D). Thus, experience-related modulation is substantially stronger in frontal compared to occipital regions.

In sum, we do not believe that the difference between visuomotor mismatch responses and playback halt responses can be explained by differences in the predictability of mismatch and playback halt events.

(2) Second, the authors admirably modified their visual-only condition to remove nausea from 6 df of movement (3D position, pitch, yaw, and roll). However, despite the fact it's far from ideal to have nauseous participants, it would appear from the figures that these modifications may have changed the responses (despite some pairwise lack of significance with small N). Specifically, the trace in S3 (6DOF) and 2E look similar - i.e., comparing the visuomotor condition to the visual condition that matches. Mismatch at 4/5 microvolts in both. Do these significantly differ from each other?

Yes, the 6DOF playback halt response shown in the previous Figure S3 and the mismatch response shown in previous Figure 2E are significantly different (Author response image 1).

Comparison of visuomotor mismatch response (A) and 6DOF playback halt response (B) from the original submission with statistics of the comparison (C).

Nevertheless, to strengthen this conclusion, we collected additional data in the 6DOF condition. We show the comparison for participants for whom both closed loop (active) and open loop sessions (6DOF) were recorded within the same recording session (14 participants) in Figure S4. Consistent with our previous findings, visuomotor mismatch responses were significantly larger than 6DOF playback halt responses (Figures S4A-S4C). And we found no evidence of a difference between 6DOF and 4DOF playback halt responses (Figures S4D and S4E).

(3) It generally seems that if the authors wish to suggest that this paradigm can be used to study prediction error responses, they need to have controlled for the actions performed and the visual events. This logic is outlined in Press, Thomas, and Yon (2023), Neurosci Biobehav Rev, and Press, Kok, and Yon (2020) Trends Cogn Sci ('learning to perceive and perceiving to learn'). For example, always requiring Ps to walk and always concurrently playing similar visual events, but modifying the extent to which the visual events can be anticipated based on action. Otherwise, it seems more accurately described as a paradigm to study the influence of action on perception, which will be generated by a number of intertwined underlying mechanisms.

We are not entirely sure we understand the point here correctly. If the reviewer is suggesting that visuomotor coupling is not describable by the ideas of predictive processing, we disagree. However, given that the papers the reviewer is pointing to are premised on what seems to be a somewhat unorthodox interpretation of predictive processing when it comes to cortical circuits, we suspect this is contributing to the misunderstanding here. Let us briefly explain. In the two papers, Press and colleagues argue that most experiments cannot distinguish between “predictive cancellation” and “gated suppression”. This is indeed relatively tricky, even when one has single neuron data. The question is, does movement simply suppress sensory feedback (as is likely the case e.g. in the famous example of the cricket), or does movement result in a precise removal of only the self-generated sensory reafference? The first good evidence of the latter happening in any system is quite recent (Keller and Hahnloser, 2009). The premise the authors build their argument on is that the theory posits that “the brain predictively ‘cancels’ expected action outcomes from perception” (from the abstract of one of the papers). This is incomplete. The minimum circuit for predictive processing is composed of 3 neuron types: positive prediction error neurons, negative prediction error neurons, and internal representation neurons. Only the positive prediction error neurons have the predictive cancellation property the authors discuss. This is not the case for either negative prediction error neurons, or for the internal representation neurons. Negative prediction error neurons are excited by predictions and suppressed by sensory input (i.e. if anything, they are “predictively amplified”). This circuit is relatively well characterized in mouse cortex – for a brief summary see (Keller and Mrsic-Flogel, 2018). Note, this is not our idea of course – the original formulation of predictive processing (Rao and Ballard, 1999) was built to explain end-stopping. These are responses to the absence of an expected line that were stronger than would be expected from classical theories (i.e. negative prediction error responses). In mouse visual cortex, we know that a sudden break in the coupling between locomotion and visual flow selectively activates layer 2/3 negative prediction error neurons. Thus, if human cortex also implements a predictive processing like circuit with positive and negative prediction error neurons, we would expect a break in visuomotor coupling to drive a measurable response in visual cortex (by exciting the population of negative prediction error neurons – this is also why we are quite excited by the phase reversal of visual and mismatch responses as this could indicate that mismatch activates negative prediction error neurons first and positive prediction error neurons later, and vice versa for visual stimulation – negative prediction error neurons are more superficial in cortex (O’Toole et al., 2023)). We do indeed find a response over occipital cortex consistent with the negative prediction error response we observe in mouse cortex. The difficulty in distinguishing “predictive cancellation” and “movement driven suppression” comes only when looking at positive prediction error type responses (that are suppressed by predictive inputs) but does not apply to negative prediction error responses. The predictive processing circuit we are testing is the one described by (Keller and Mrsic-Flogel, 2018; Rao and Ballard, 1999), and here the break in visuomotor coupling is a stimulus that drives negative prediction error responses. Note, other authors who have thought about cortical implementations of predictive processing (e.g. (Bastos et al., 2012)) have glossed over the problem that individual neurons cannot trivially encode both positive and negative errors. Prediction errors are a signed quantity. If neurons signal prediction errors in firing rates and are close to zero firing rate at baseline (as is the case in layer 2/3 of cortex), they cannot (short of rather exotic ideas) encode a signed prediction error. Hence such proposals are not very useful for thinking about prediction error responses in cortex. For these reasons, we see no problem with referring to the response as a prediction error response. This is in line with a large body of mouse research (using a nearly identical paradigm) on the topic.

One could of course argue that gated suppression could also mean that movement relieves suppression. Thus, one could assume that some neurons are suppressed by movement while others are enhanced. If one allows for enough neuron and stimulus specificity in the precision of the movement related suppression and enhancement of responses, the two models (predictive processing and gated suppression) become equivalent, and the discussion becomes semantic. See (Vasilevskaya et al., 2023) for an extended discussion on this point, and the reasons why we think predictive processing is a more useful model than gated suppression (keep in mind, gated suppression only explains the data if we allow for stimulus/neuron specific gain factors of the suppression, in which case the two models are equivalent).

More minor points:

(1) I was also wondering whether the authors may consider the findings in frontal electrodes more closely. Within the statistical tests of the frontal electrodes against 0, as displayed in Figure 3c, the insignificance of the effect of Fp2 seems attributable to the small included sample size of just 13 participants for this electrode, as listed in Table S1, in combination with a single outlier skewing the result. The small sample size stands out especially in comparison to the sample size at occipital electrodes, which is double and therefore enjoys far more statistical power. It looks like the selected time window is not perfectly aligned for determining a frontal effect, and also the distribution in 3B looks like responses are absent in more central electrodes but present in occipital and frontal ones. I realise the focus of analysis is on visual processing, but there are likely to be researchers who find the frontal effect just as interesting.

That is correct; our data in frontal electrodes was likely underpowered. The reason we have fewer data in frontal electrodes is that eye-blink artifacts are particularly strong in frontal channels, resulting in a larger proportion of trials failing to meet our data inclusion criteria. We have now added more data from frontal and occipital electrodes by including additional experimental sessions. In addition, we applied less stringent trial-exclusion criteria, requiring that no artifacts occur within the time window −0.5 to 1 s relative to the event trigger (instead of −0.5 to 2 s). This adjustment allowed us to retain a larger number of trials. As anticipated by the reviewer, this increase in data was sufficient to confirm a significant response to the visuomotor mismatch event at both frontal electrodes (Figure 3C). The expanded dataset also revealed a significant difference in response onset times between occipital and frontal electrodes (Figure 3E), an effect that was not significant previously. In addition, we have included analysis comparing early and late mismatch responses in frontal and occipital electrodes (Figure 4).

(2) It is claimed throughout the manuscript that the 'strongest predictor (of sensory input) - by consistency of coupling - is self-generated movement'. This claim is going to be hard to validate, and I wonder whether it might be received better by the community to be framed as an especially strong predictor rather than necessarily the strongest. If I hear an ambulance siren, this is an especially strong predictor of subsequent visual events. If I see a traffic light turn red, then yellow, I can be pretty certain what will happen next. Etc.

This is a statistical argument. Every movement – throughout life – is directly and immediately coupled to sensory feedback and has been throughout evolutionary history. The vast majority of visual input you receive (we estimate, well above 99%) is the consequence of your own movements (e.g. every few 100 ms your eye movements cause a full field change in your visual input). The same is likely true of proprioceptive and somatosensory input – the vast majority is the direct consequence of your own movements (not other people poking you). This is likely different in the auditory system where a much larger fraction of the input is externally driven (depending a bit on how much one likes to talk). But even here the best predictor is self-motion (most non-self-generated sounds one experiences in life are very difficult to predict with millisecond precision). The example the reviewer gives is a good illustration of this. Take the siren that hails the appearance of an ambulance. The siren tells us that an ambulance will appear, but not how it will look, not when exactly it will appear, and with only very low resolution as to where it will appear. Incidentally, if you ask people to draw an ambulance they tend to draw a WWII style white square vehicle with a red cross on the side – a style of ambulance they likely have not ever seen in life. Their visual predictions of what they are about to see are very low resolution. We catastrophically fail at making pixel perfect predictions from learned stimulus associations of this nature. The traffic light example is difficult to compare to visual feedback control of movement as it is a much simpler prediction of a single bit in the form of a change in color of an existing object.

In addition, consider how often (in life) you have seen an ambulance after hearing it? 100 times maybe? Maybe less. How often have you seen traffic lights change - 10 000 times? 100 000 times? Now consider, how often you have experienced the visual consequences of moving your head or eyes to the left (keep in mind this includes micro saccades) – at a conservative, once per second, that is somewhere on the order of 1 000 000 000. This is not even in the same ballpark. Our brains can certainly learn to make the ambulance and traffic light type predictions - to some extent - but by far the best predictor of sensory feedback (simply by virtue of the physics of how our body interacts with the world) is self-motion.

We think this is an argument we can make based on first principles, and one that is frequently overlooked in the field, as experiments often focus on training people or animals to learn novel associations that, especially in the case of mice, we often have no idea whether cortical circuits can even learn. We should focus experiments on the predictive systems our brains have evolved since long before the evolutionary appearance of ambulances and traffic lights. We understand that the reviewer may disagree with this, but unless the reviewer has a concrete example of an even stronger predictor (as measured by frequency of experience, consistency in coupling, and precision in timing – we can’t think of one), it is a point we will make.

(3) The checkerboard inversion response at 48 ms is incredibly rapid. Can the authors comment more on what may drive this exceptionally fast response? It was my understanding that responses in this time window can only be isolated with human EEG by presenting spatially polarized events (cf. c1, e.g., Alilovic, Timmermans, Reteig, van Gaal, Slagter, 2019, Cerebral Cortex).

We don’t know, but it is not inconsistent with previous reports. For example, compare the “standing” and “fast walking” target ERP responses in Figure 5 of (Gramann et al., 2010). Both here and in our data, the fast response peak is only really apparent in the direct comparison of visual responses recorded while participants were walking to those when they were stationary.

While we have taken great care to calibrate the timing of the visual display with the EEG recording, one could be worried that the alignment is off by as much as tens of milliseconds. However, even if this were so, one could use P1 as a reference and determine that the fast peak roughly precedes P1 by about 40 ms. Which again would result in a latency of about 50 ms of the fast walking peak (assuming P1 peaks at about 90 ms). In sum, we have added a reference to the previous work (that we found thanks to the reviewer’s comment) but fear we have nothing intelligent to say beyond that.

Reviewer #2 (Public review):

Summary:

This study investigates whether visuomotor mismatch responses can be detected in humans. By adapting paradigms from rodent studies, the authors report EEG evidence of mismatch responses during visuomotor conditions and compare them to visual-only stimulation and mismatch responses in other modalities.

Strengths:

(1) The authors use a creative experimental design to elicit visuomotor mismatch responses in humans.

(2) The study provides an initial dataset and analytical framework that could support future research on human visuomotor prediction errors.

Weaknesses:

(1) Methodological issues (e.g., volume conduction, channel selection, lack of control for eye movements) make it difficult to confidently attribute the observed mismatch responses to activity in visual cortical regions.

(2) A very large portion of the data was excluded due to motion artefacts, raising concerns about statistical power and representativeness. The criteria for trial inclusion and the number of accepted trials per participant appear arbitrary and not justified with reference to EEG reliability standards.

(3) The comparison across sensory modalities (e.g., auditory vs. visual mismatch responses) is conceptually interesting, but due to the choice of analyzing auditory mismatch responses over occipital channels, it has limited interpretability.

We have responded to these points in the more detailed itemization below.

The authors successfully demonstrate that visuomotor mismatch paradigms can, in principle, be applied in human EEG. However, due to the issues outlined above, the current findings are relatively preliminary. If validated with improved methodology, this approach could significantly advance our understanding of predictive processing in the human visual system and provide a translational bridge between rodent and human work.

Reviewer #2 (Recommendations for the authors):

Overall, the study addresses an interesting and underexplored question (translation of the visuomotor mismatch responses observed in rodents to humans). Below, please find a list of specific suggestions for improvement

Introduction:

(1) "updating internal representations and internal models" - what is the difference between the two, and why is it relevant to this study?

In a nutshell, an internal model is the synaptic weight matrix that transforms between coding spaces. An internal representation is the activity pattern coding for the current representation. See (Aizenbud et al., 2025; Keller and Mrsic-Flogel, 2018) for more lengthy elaborations. The fact that the mechanism used for representation update can also be used to update internal models (i.e. solve the credit assignment problem) is likely the prime advantage of predictive processing (see work from the Bogacz lab). The relevance to the current study is justifying why predictive processing is a reasonable hypothesis for the function of cortex.

(2) "Certain stimuli can be predicted from the preceding sensory input" vs. "Predictions can also be based on memory" - how are these two different? Do you mean specific (e.g., long-term associative or episodic) memory types in the latter?

Correct, this is an arbitrary distinction that primarily makes sense in the light of experimental approaches. In this particular case, we were talking about spatial memory. We made this explicit to increase clarity.

(3) "the strongest predictor - by consistency of coupling - is self-generated movement"

(a) Externally induced movement, while not self-generated and therefore not predicted, will also generate sensory coupling, so is it really only about consistency?

Externally induced movement (as in somebody else moving one’s arm we are not sure this is what the reviewer means) will induce sensory-sensory coupling but not sensorimotor coupling. We might be misunderstanding the point. In case the reviewer means stimuli that trigger movement as in us asking participants to walk, or a sudden startle stimulus that makes them jump in all such cases there are of course sensorimotor predictions. Sensorimotor predictions are driven by efference copies of the motor command thus all movements whether ‘voluntarily’ executed or triggered by an external stimulus will drive sensorimotor predictions. (All of this of course assumes that the predictive processing theory is correct.)

(b) Do you mean temporal consistency (minimal lags), statistical contingencies (same movements linked to the same sensory inputs), or both? How does it differentiate sensorimotor/visuomotor mismatch responses from responses to incongruent stimuli in sensory modalities (e.g. audiovisual)?

Both. We have rephrased the sentence to try to make this clearer. See also response to reviewer 1 minor point 2 above.

How does it differentiate sensorimotor/visuomotor mismatch responses from responses to incongruent stimuli in sensory modalities (e.g. audiovisual)?

Most cross-modal associations are much less consistent (the exact sound of a glass shattering is always slightly different and impossible for us to predict), and orders of magnitude less frequently experienced, than sensorimotor associations. Again, see also response to reviewer 1 minor point 2 above.

(4) "Every movement is directly coupled to sensory feedback throughout life"

This may be the case for proprioceptive and/or somatosensory feedback, but not necessarily for visual feedback (e.g., a mouse moving its tail), which is the topic of the study.

Correct, there are movements that can be disconnected from visual feedback. Most of the time, most movements however are not, and we are studying one of the more prominent ones that is clearly not decoupled locomotion. The contrast we aim to highlight here very prominently is that there is still this vague idea in the field that you can take a participant, or a mouse, and expose them/it to a few tens or hundreds of trials of some sensory stimulus contingency and then probe for prediction error responses to a pattern only recently if at all learned. Given the life-long experience of subjects and mice, is it really surprising that oddball responses are less strong than a sensorimotor mismatch?

(5) "However, the overall level of this motor-related activity is much higher than one would expect simply from predictions of visual feedback that are compared against visual input."

Could you please clarify what one would expect in this case, and/or back it up with citations?

This is in reference to the fact that there are very strong movement related signals in the mouse visual cortex that persist even when the mouse is in complete darkness. In darkness, movements should not trigger any visual feedback change hence the activity is difficult to explain as a movement related prediction of visual flow. We have rephrased this section of the introduction to make this clearer.

(6) "The more precise the prediction and comparison, the less motor-related activity should be detectable in visual cortex."

I think this conflates two issues. A good match between prediction and input would indeed result in sensory attenuation. However, sensory precision, at least in active inference, can upregulate prediction error responses. Since predictions cannot be assumed to be perfect (due to external or internal noise), increased precision may therefore augment activity. See e.g. https://doi.org/10.1007/s10339-013-0571-3

We agree with the reviewer – the phrasing here was misleading. We do not mean precision in the predictive processing sense, but the precision of sensorimotor control necessary for the behavior. We have rephrased the corresponding section of the manuscript.

(7) Neither the introduction nor the discussion refers to previous human EEG studies on sensorimotor mismatch responses, where sensory feedback doesn't match motor actions (e.g. https://doi.org/10.3758/s13423-021-01992-z ; https://www.sciencedirect.com/science/article/pii/S0028393214003777 ; https://www.sciencedirect.com/science/article/pii/S0028393219301265).

The studies cited by the reviewer primarily test how discrete violations of learned action–outcome associations are represented in the brain, whereas our visuomotor mismatch paradigm probes violations of continuous sensorimotor coupling during ongoing action. The paradigms are conceptually different both in how strong the coupling is (lifelong vs. learned in the experiment), and in how prediction errors are likely used (visuomotor control vs. stimulus detection). We have added a brief part to our introduction discussing this.

Results:

(1) A very large proportion of the dataset was excluded due to movement artefacts. This is rather problematic as

(a) the rationale behind finding mismatch responses is that motion-related (neural) signals should affect visual cortical activity, so it's essential to disentangle these neural signals from artefacts;

Correct, we excluded 21.7% of the total data for visuomotor mismatch paradigm. Note, this percentage compares to other similar studies of EEG recordings during movement (Oliveira et al., 2016). By “problematic”, we assume the reviewer means the fact that we have artefacts, not that we exclude trials with artefacts. The movement artefacts are typically caused by the acceleration during stepping in participants with a heavy gait. None of these movement artefacts are time locked to any of the responses we investigate. Thus, they should just appear as increased levels of noise if not excluded. We don’t understand why the reviewer thinks this is particularly problematic for our analysis/conclusions (beyond the trivial consequence of increasing noise levels that would only cause us to underestimate the strength of the mismatch signals we report).

(b) the criterion for the number of trials of 15 triggers (per condition?) is arbitrary and lower than widely used in the literature, so authors should demonstrate that this is a sufficient number to observe a measurable ERP even for those participants with 15 triggers;

We have between 16 and 25 visuomotor mismatch events per participant. Author response image 2 is a selection of single participant examples with different number of trials. The number of mismatch events is limited by the fact that we introduce them approximately every 10 - 15 s and have a total duration of the closed loop session of 5 minutes. Thus, on average, we expect to have 24 mismatch events. But we are not sure we understand the logic of the comment, if we set exclusion too low, we just risk losing a response in the noise. And we clearly have stronger and higher signal to noise mismatch responses with an average of 20 trials compared to visual responses during movement with an average of 40 trials or MMN responses with an average of 28 trials.

Reliable ERPs can be observed with as few as 16 trials across EEG channels.

(A) Histograms showing the distribution of the number of valid mismatch trials per participant for each electrode pair (Fp1–2, C3–4, P3–4, O1–2). (B) Representative EEG responses to visuomotor mismatch events from a single participant, recorded at electrode pairs Fp1–2, C3–4, P3–4, and O1–2. Waveforms were computed using the indicated number of trials (shown above each trace). Dashed vertical red lines are onset and offset of the visuomotor mismatch.

(c) it seems that the seemingly static "visual" condition resulted in a larger proportion of data rejected due to movement (or, as later mentioned, nausea) than the "visuomotor" condition, which is counterintuitive and needs further explanation;

This is a misunderstanding the ‘visual paradigm’ the reviewer is referring to are the experiments shown in Figure 1. Here we record visual responses in both sitting and walking participants. In this experiment, as in others, exclusion was primarily driven by part of the paradigm where the subjects were moving. To make this clearer we have added Table S2 to the manuscript that provides an overview of trials excluded by paradigm and session.

(d) authors mention eye movements as a potential issue, which should be possible to detect from frontal channels. Additionally, it's not entirely clear how many datasets were discarded (the results section mentions 19/48 in the visual condition, then 4+11 in the playback condition - isn't this the same condition?)

The visual paradigm corresponds to the data shown in Figure 1, in which participants viewed a flipping checkerboard in both a walking and a stationary session. The open loop session is part of the visuomotor paradigm shown in Figure 2, where participants were exposed to a replay of the visual flow that had been self-generated during the preceding closed loop session, including the visual flow halts that constituted visuomotor mismatches in the closed loop session. Please note, to avoid such confusion, we have attempted to standardize the usage of paradigm (visual vs. visuomotor) and session (sitting vs. walking, and closed loop vs. open loop) throughout. In addition, we have added a table to summarize the number of excluded trials by paradigm and session as Table S2 to the manuscript.

In comments 1 and 2 of the public review, the reviewer also points out that we did not control for eye movements and we presume relatedly claims that we did not use common EEG reliability standards. Regarding the first point, we performed additional experiments in an independent cohort of participants to test whether eye movements could account for the visuomotor mismatch responses. We recorded eye movements during closed loop sessions and found that changes in eye speed (Figure S7A) or blink rate (Figure S7B) following the mismatch stimulus had a longer latency than visuomotor mismatch responses in EEG. This suggests that the visuomotor mismatch response cannot be explained by eye blinks or changes in eye movement speed. Regarding the second point, we are not sure we understand. Trial exclusion based on a fixed voltage threshold of 100 µV is relatively common, and our rejection rates are on par, and particularly on occipital electrodes even lower, with other work in EEG recordings during locomotion or movement (see e.g. (Oliveira et al., 2016)).

Nevertheless, we did attempt to apply independent component analysis (ICA) based filtering to the EEG data (Delorme and Makeig, 2004). However, these methods were designed for high channel density recordings. With only 8 channels, ICA is unable to reliably isolate eye movement or motion artefact components of the EEG. To illustrate this, we tested two artifact-rejection strategies. In the first approach, components associated with non-neural artifacts (e.g., muscle activity, line noise, eye movements) were removed only if at least 90% of the component’s variance was assigned to a single artifact class (Author response image 3A). In the second, more permissive approach aimed specifically at reducing eye movement artifacts, components were removed if artifact-related activity exceeded 90% for non-eye artifacts, while the threshold for eye-related components was lowered to 60% (Author response image 3C). We lowered the threshold for excluding eye-related components to ensure that EEG signals influenced by eye movements were effectively removed. In both cases - whether the eye-component threshold was set to 90% or 60% - the averaged responses to visuomotor mismatch trials remained largely similar to the previously reported data, despite higher noise in some traces. Interestingly, when we then followed the ICA filtering by our voltage threshold based exclusion with a threshold of 100 µV, the resulting traces closely resembled the patterns described in the paper (Author response image 3B and 3D). Thus, we conclude the nonICA filtered responses are easier to interpret, free of any potential ICA filtering artifacts, and far less parameter choice (of the ICA filtering) dependent.

Removal of artifacts identified with ICA does not change the visuomotor mismatch responses.

(A) Visuomotor mismatch responses recorded from occipital electrodes after artifact correction. Components associated with non-neural artifacts (e.g., muscle activity, line noise, eye movements) were removed only if ≥90% of the component’s variance was attributed to a single artifact class. Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (B) As in A, but excluding trials with amplitudes exceeding 100 µV. (C) As in A, but components were removed if artifact-related activity exceeded 90% for non-ocular artifacts, while the threshold for eye-related components was lowered to 60%. (D) As in C, but excluding trials with amplitudes exceeding 100 µV.

(2) The finding that mismatch responses are observed at all channels, with differences in amplitudes but not latencies, indicates that volume conduction may affect the results. I would strongly suggest accounting for this using a method appropriate for the very small number of channels, e.g., phase lag index.

We are not sure we understand. The phase lag index is a method to estimate functional connectivity in a way that corrects for volume conduction (using phase lag). We make no claims about functional connectivity; thus, we are not sure what the reviewer is suggesting we do. The fact that the visual and visuomotor mismatch responses were measurable on all electrodes could indeed be in part explained by volume conduction, but we see no way to estimate the volume conduction contribution. From mouse calcium imaging data, we know that both visual and visuomotor mismatch responses spread across large parts of dorsal cortex (including frontal regions like the ACC).

With the addition of new data, the latency difference between occipital and frontal electrodes - previously observed only as a trend - is now statistically significant (Figure 3E). Occipital responses emerge earlier than frontal responses, suggesting that mismatch-related activity likely originates in sensory visual regions and subsequently propagates to more frontal areas, as similar to what had been reported in mouse cortex (Heindorf and Keller, 2024).

(3) The authors compare different types of mismatch responses (including auditory oddballs) in the same set of (occipital) channels, but doesn't this undermine the spatial specificity of the results? Classical auditory mismatch negativity is typically observed over central channels, so weaker amplitudes of auditory mismatch responses in occipital channels are likely trivially explained by modality differences. As such, I'm not convinced that this comparison is informative even in a qualitative manner.

To address this point, we conducted additional auditory oddball experiments with recordings over the auditory cortex (channels T3, T4, T5, and T6). Given our central reference, these channels should capture the strongest mismatch negativity. The amplitude of the visuomotor mismatch response exceeded that of mismatch negativity on all tested channels (new Figures S8 and S9).

(4) On a similar note, is the polarity reversal found for visual vs. mismatch responses specific to occipital channels?

Thank you for this interesting question. In fact, polarity reversal was consistently observed across all recorded channels; this has now been added as a main figure to the manuscript (Figure 5).

(5) Figure S4C seems to cut off one outlier, and I don't see this outlier included in the boxplot.

Correct, that is why we describe the boxplots in the figure legend as: “Boxes mark median, quartiles, and range of data not considered outliers.” The axes were now adjusted to include all data points.

Discussion:

"A central tenet of the cortical circuit for predictive processing is the split into separate populations of neurons that compute positive and negative prediction errors (Keller and Mrsic-Flogel, 2018; Rao and Ballard, 1999)" - this may be the case for visuomotor mismatch signals or reward prediction errors, but signed PEs do not play a central role in other proposed microcircuits for predictive processing in the perceptual domain (e.g. Bastos)

Signed prediction errors do not play a central role in proposed cortical microcircuits for predictive processing that do not burden themselves with making a concrete proposal for the implementation of the prediction error computation. The (Bastos et al., 2012) work is a good example of this. The equation for the error term provided in that paper is clearly signed (nothing stops the error from going negative), but no proposal is made for how layer 2/3 excitatory neurons are supposed to signal this quantity. With baseline activity levels close to zero in layer 2/3, there really is only one way to do this, and that is separate populations of negative and positive prediction error neurons. With non-zero baseline firing rate, one could do this bidirectionally around a mean firing rate (as is typically thought of dopaminergic RPE neurons). There are more abstract Bayesian implementations that assume logarithmic transformations that could also implement a prediction error-like system without negative firing rates. But given the absence of any physiological evidence, we will refrain from discussing these. However, most importantly, there is now considerable evidence for the existence of both negative and positive prediction error neurons in layer 2/3 of mouse visual cortex. Thus, by “cortical circuit for predictive processing” we here mean those that make biologically plausible proposals for prediction error computations. Also note, the (Rao and Ballard, 1999) model is probably the prime example for what the reviewer calls a proposed microcircuit for predictive processing in the “perceptual domain”.

Reviewer #3 (Public review):

Summary:

Solyga, Zelechowski, and Keller present a concise report of an innovative study demonstrating clear visuomotor mismatch responses in ambulating humans, using a mobile EEG setup and virtual reality. Human subjects walked around a virtual corridor while EEGs were recorded. Occasionally, motion and visual flow were uncoupled, and this evoked a mismatch response that was strongest in occipitally placed electrodes and had a considerable signal-to-noise ratio. It was robust across participants and could not be explained by the visual stimulus alone.

Strengths:

This is an important extension of their prior work in mice, and represents an elegant translation of those previous findings to humans, where future work can inform theories of e.g., psychiatric diseases that are believed to involve disordered predictive processing. For the most part, the authors are appropriately circumspect in their interpretations and discussions of the implications. I found the discussion of the polarity differences they found in light of separate positive and negative prediction errors, intriguing.

Weaknesses:

The primary weaknesses rest in how the results are sold and interpreted.

Most notably, the interpretation of the results of the comparison of visuomotor mismatches to the passive auditory oddball induced mismatch responses is inappropriate, as suboptimal electrode choices, unclear matching of trial numbers, and other factors. To clarify, regarding the auditory oddball portion in Figure 5, the data quality is a concern for the auditory ERPs, and the choice of Occipital electrodes is a likely culprit. Typically, auditory evoked responses are maximal at Cz or FCz, although these contacts don't seem to be available with this setup. In general, caution is warranted in comparing ERP peaks between two different sensory modalities - especially if attention is directed elsewhere (to a silent movie) during one recording and not during the other. The authors discuss this as a purely "qualitative" comparison in the text, which is appreciated, and do acknowledge the limitations within the results section, but the figure title and, importantly, the abstract set a different tone. At least, for comparisons between auditory mismatch and visuomotor mismatch, trial numbers need to be equated, as ERP magnitude can be augmented by noise (which reduces with increased numbers of trials in the average).

To address this point, we conducted additional auditory oddball experiments with recordings over the auditory cortex (channels T3, T4, T5, and T6). Given our central reference, these channels should capture the strongest mismatch negativity. Nevertheless, the amplitude of the visuomotor mismatch response exceeded that of mismatch negativity on all tested channels (these results are now shown in the new Figures S8 and S9), and the response power was significantly greater for the visuomotor mismatch than for mismatch negativity. Independent of electrode we test, the visuomotor mismatch response has a power 5 to 10 times higher than that of the MMN response. And the number of trials per participant that met quality criteria was comparable between the visuomotor mismatch paradigm (mean = 23 trials) and the auditory mismatch paradigm (mean = 28 trials) (Author response image 4).

Number of trials included for analysis is comparable between visuomotor and oddball paradigm.

(A) Histogram showing the distribution of the number of valid trials per participant for O1-2 electrode pair in visuomotor mismatch paradigm. (B) Same as in A but for deviant stimulus presentations in the oddball paradigm.

And more generally, the size of the mismatch event at the scalp does not scale one-to-one with the size at the level of the neural tissue. One can imagine a number of variables that impact scalp level magnitudes, which are orthogonal to actual cortex-level activation - the size, spread, and polarity variance of the activated source (which all would diminish amplitude at the scalp due to polyphasic summation/cancelation). The variance of phase to a stimulus across trials (cross trial phase locking) vs magnitude of underlying power - the former, in theory, relates to bottom-up activity and the latter can reflect feedback (which has more variability in time across trials; the distance of the scalp electrode from the activated tissue (which, for the auditory system, would be larger (FCz to superior temporal gyrus) than for the visual system (O1 to V1/2)). None of this precludes the inclusion of the auditory mismatch, which is a strength of the study, but interpretations about this supporting a supremacy of sensory-motor mismatch - regardless of validity - are not warranted. I would recommend changing the way this is presented in the abstract.

We agree with the point that the EEG response does not need to reflect the total cortical activation. However, the discussion in the abstract (and elsewhere) is in the context of clinical experiments where the underlying cortical activity pattern is irrelevant if it does not trigger a clinically measurable (by EEG in this case) response. The abstract only makes a comparison to MMN implicitly in this sentence “Second, a paradigm that can trigger strong prediction error responses and consequently requires shorter recording times could simplify experiments in a clinical setting.” We are not sure how to phrase this even more carefully – the statement at face value is a truism. The reviewer, we assume, takes exception to the unstated implication that visuomotor prediction errors trigger stronger responses than MMN. Given the data we have, we assume most authors would not consider it an overstatement to make that claim outright.

Otherwise, the data are of adequate quality to derive most of their conclusions.

The authors claim that the mismatch responses emanate from within the occipital cortex, but I would require denser scalp coverage or a demonstration of consistent impedances across electrodes and across subjects to make conclusions about the underlying cortical sources (especially given the latencies of their peaks). In EEG, the distribution of voltage on the scalp is, of course, related to but not directly reflective of the distribution of the underlying sources. The authors are mostly careful in their discussion of this, but I would strongly recommend changing the work choice of "in occipital cortex" to "over occipital cortex" or even "posteriorly distributed". Even with very dense electrode coverage and co-registration to MRIs for the generation of forward models that constrain solutions, source localization of EEG signals is very challenging and not a simple problem. Given the convoluted and interior nature of human V1, the ability to reliably detect early evoked responses (which show the mismatch in mouse models) at the scalp in ERP peaks is challenging - especially if one is collapsing ERPs across subjects. And - given the latency of the mismatch responses, I'd imagine that many distributed cortical regions contribute to the responses seen at the scalp.

This is an excellent point we have rephrased throughout to “over occipital cortex” instead of “in occipital cortex”.

I think that Figure 3C, but as a difference of visual mismatch vs halting flow alone (in the open loop) might be additionally informative, as it clarifies exactly where the pure "mismatch" or prediction error is represented.

We performed the analysis as suggested (Author response image 5). Visuomotor mismatch responses are stronger on all electrodes compared to playback halt responses. This difference is also larger in data recorded on occipital electrodes.

Comparison of the difference between visuomotor mismatch and playback halt on all electrodes.

Average response strength was calculated within a 100 ms window centered on the peak of the average visuomotor mismatch response across all electrodes. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from one participant. **: p<0.01, *: p<0.05, Fp1-2: 20 participants, C3-4: 31 participants, P3-4: 35 participants, O1-2: 32 participants.

As a suggestion, the authors are encouraged to analyse time-frequency power and phase locking for these mismatch responses, as is common in much of the literature (see Roach et al 2008, Schizophrenia Bulletin). This is not to say that doing so will yield insights into oscillations per se, but converting the data to the time-frequency domain provides another perspective that has some advantages. It fosters translations to rodent models, as ERP peaks do not map well between species, but e.g., delta-theta power does (see Lee et al 2018, Neuropsychopharmacology; Javitt et al 2018, Schizophrenia research; Gallimore et al 2023, Cereb Ctx). Further, ERP peaks can be influenced by the actual neuroanatomy of an individual (especially for quantifying V1 responses). Time frequency analyses may aid in interpreting the "early negative deflection with a peak latency of 48 ms " finding as well.

We have performed time–frequency power and phase-locking analyses for both visual responses (Author response image 6 and Author response image 7) and visuomotor mismatch and playback halt responses (Author response image 8 and Author response image 9), as suggested. We have added the results of these analyses here, as these are not fully developed yet. We may add these to a future publication, for which we would properly want to quantify stability of these effects.

In brief, time–frequency representations of power did identify potentially interesting differences between walking and sitting sessions in the visual paradigm. Inter-trial phase coherence (ITPC) revealed an early increase in alpha-band synchronization suggesting that phase alignment of alpha oscillations may contribute to the early differences in visual responses between walking and sitting. The same analyses were applied to visuomotor mismatch and playback halt responses. Time–frequency power analysis revealed an increase in delta-band power during visuomotor mismatch, consistent with previous reports linking delta activity to prediction error processing, including reward prediction errors (Cavanagh, 2015), unexpected final words (Webb and Sohoglu, 2025), and visual deviance detection (West et al., 2024). Notably, it appears as if the increase in delta power emerged first over occipital electrodes and appeared later over more frontal electrodes, forming a spatiotemporal gradient of onset across the scalp.

Delta power changes were markedly reduced in the playback halt responses at the time of visual flow cessation. While some power changes were observed, they occurred primarily at visual flow onset rather than at flow offset. Inter-trial phase coherence analysis further revealed delta-band synchronization over occipital electrodes following visuomotor mismatch, whereas the playback halt response showed strong phase synchronization in both delta and theta bands following visual flow onset.

Time–frequency representations of EEG power changes during the visual paradigm.

(A) Time–frequency maps showing changes in spectral power relative to baseline for electrodes Fp1–2, C3–4, P3–4, and O1–2 following checkerboard reversal in the sitting session. The dashed red vertical line indicates the time of the checkerboard reversal (0 s). (B) As in A, but recorded while participants were walking.

Inter-trial phase coherence (ITPC) for visual trials during sitting and walking.

(A) ITPC across trials for electrode pairs Fp1–2, C3–4, P3–4, and O1–2 following checkerboard reversal in the sitting session. The dashed red vertical line marks the time of the checkerboard reversal (0 s). (B) As in A, but recorded during walking.

Time–frequency representations of EEG power changes during visuomotor mismatch and playback halt responses.

(A) Time–frequency maps showing changes in spectral power relative to baseline for electrodes Fp1–2, C3–4, P3–4, and O1–2 following visuomotor mismatch presentation. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (B) As in A, but for playback halts.

Inter-trial phase coherence (ITPC) for the visuomotor mismatch and playback halt responses.

(A) ITPC across trials for electrode pairs Fp1–2, C3–4, P3–4, and O1–2 following visuomotor mismatch presentation. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (B) As in A, but for playback halts.

Finally, the sentence in the abstract that this paradigm " can trigger strong prediction error responses and consequently requires shorter recording times would simplify experiments in a clinical setting" is a nice setup to the paper, but the very fact that one third of recordings had to be removed due to movement artifact, and that hairstyle modulates the recording SnR, is reason that this paradigm, using the reported equipment, may have limited clinical utility in its current form. Further, auditory oddball paradigms are of great clinical utility because they do not require explicit attention and can be recorded very quickly with no behavioral involvement of a hospitalized patient. This should be discussed, although it does not detract from the overall scientific importance of the study. The authors should reconsider putting this statement in the abstract.

We have added a paragraph to the discussion to address these points. Note, we get robust and strong responses with very few trials (Author response image 2). The fact that we need to discard up to 21.7 % of trials due to movement/eye blink artefacts, does little to change the fact that we need much fewer trials and have larger and more robust responses compared to other EEG paradigms. Finally, we understand that sometimes not needing participants to pay attention to the task is useful. However, having a paradigm that is engaging and fun for participants and takes 5 minutes of recording time is probably equally often of advantage.

Reviewer #3 (Recommendations for the authors):

Minor points:

(1) In the Introduction, I'm not sure that the logic comes through as to what the authors aim to illustrate by comparing mice to humans, in terms of precision and "movement modulation". In some cases, the precision of the comparison is referred to, and in others, the precision of the prediction (I think?). I'm not sure if they mean for this to be different or not. Simlarly, on line 81, "If indeed the precision of visuomotor coupling determines the amount of motor modulation of visual responses" - here I'm a little confused, as "amount of motor modulation" to me, the term "modulation" refers to a conditional modifier (if moving, than suppress visual movement resposnes. if not moving, then amplify visual movement repssones) rather than movement driven activity. The way I'm reading it, the authors mean the latter, but I could be misunderstanding.

We have rephrased this section of the introduction.

(2) I think it could be helpful, in the sentence starting on line 65, to reiterate that this observation of higher-than-expected motor activity in V1 is in mice (if I'm understanding it correctly). I also found myself tangled up in the difference between motor-related activity in V1 and motor-modulation in V1 in this paragraph.

We have rephrased this section of the introduction.

(3) For signal power, was the amplitude squared on individual trials prior to averaging, or after averaging? If prior, it would help with separating amplitude modulations from phase variance.

In our previous analysis, power was computed by squaring the amplitude after trial averaging (Author response image 10A). We repeated the analysis using the alternative approach in which power was calculated for individual trials and then averaged (Author response image 10B). Although this method yields substantially higher absolute power values, the overall pattern of results remains unchanged: visuomotor mismatch responses continue to show significantly higher power than visual responses. To look at the phase variance we additionally analyze inter-trial phase coherence (Author response image 7 and Author response image 9).

Visuomotor mismatch responses have more power compared to visual responses.

(A) Comparison of power between visuomotor mismatch and visual responses, calculated within a 0 - 0.5 s time window following stimulus onset. Power was computed by squaring the amplitude after trial averaging. Boxes indicate the median and interquartile range, with whiskers showing the range excluding outliers; circles represent data from individual participants. ***p < 0.001. (B) Same comparison as in (A), but with power calculated by squaring the amplitude of individual trials prior to averaging.

(4) The "the world suddenly flew forward!" response from the participant, I understand, and I believe that it is useful to illustrate a point. I do not understand the "Are you printing this? - Hi Mom! " part of the participant response, and I'm not sure it adds to the paper, beyond amusement, which seems inappropriate.

One of the authors (the one who did none of the experiments) finds this endlessly hilarious and as the reviewer notes, it might add amusement more generally. “Inappropriate” might be a bit harsh – according to our favorite AI chatbot: “Amusement provides significant mental, physical, and social value by offering a necessary escape from routine, reducing stress, and fostering a connection. It enhances well-being through endorphin-releasing experiences and encourages social bonding, learning, and joy.” Nevertheless, we have censored the offending passage.

Aizenbud, I., Audette, N., Auksztulewicz, R., Basiński, K., Bastos, A.M., Berry, M., Canales-Johnson, A., Choi, H., Clopath, C., Cohen, U., Costa, R.P., Filippo, R.D., Doronin, R., Errington, S.P., Gavornik, J.P., Gillon, C.J., Granier, A., Hamm, J.P., Hertäg, L., Kennedy, H., Kumar, S., Ladd, A., Ladret, H., Lecoq, J.A., Maier, A., McCarthy, P., Mei, J., Mejias, J., Mikulasch, F., Mudrik, N., Najafi, F., Nejad, K., Nejat, H., Oweiss, K., Petrovici, M.A., Priesemann, V., Rudelt, L., Ruediger, S., Russo, S., Salatiello, A., Senn, W., Sennesh, E., Sima, S., Uran, C., Vasilevskaya, A., Vezoli, J., Vinck, M., Westerberg, J.A., Wilmes, K., Xiong, Y.S., 2025. Neural mechanisms of predictive processing: a collaborative community experiment through the OpenScope program. https://doi.org/10.48550/arXiv.2504.09614

Bastos, A.M., Usrey, W.M., Adams, R.A., Mangun, G.R., Fries, P., Friston, K.J., 2012. Canonical microcircuits for predictive coding. Neuron 76, 695–711. https://doi.org/10.1016/j.neuron.2012.10.038

Cavanagh, J.F., 2015. Cortical delta activity reflects reward prediction error and related behavioral adjustments, but at different times. NeuroImage 110, 205–216. https://doi.org/10.1016/j.neuroimage.2015.02.007

Delorme, A., Makeig, S., 2004. EEGLAB: an open source toolbox for analysis of single-trial EEG dynamics including independent component analysis. J. Neurosci. Methods 134, 9–21. https://doi.org/10.1016/j.jneumeth.2003.10.009

Gramann, K., Gwin, J.T., Bigdely-Shamlo, N., Ferris, D.P., Makeig, S., 2010. Visual evoked responses during standing and walking. Front. Hum. Neurosci. 4, 202. https://doi.org/10.3389/fnhum.2010.00202

Heindorf, M., Keller, G.B., 2024. Antipsychotic drugs selectively decorrelate long-range interactions in deep cortical layers. eLife 12, RP86805. https://doi.org/10.7554/eLife.86805

Keller, G.B., Hahnloser, R.H.R., 2009. Neural processing of auditory feedback during vocal practice in a songbird. Nature 457, 187–90. https://doi.org/10.1038/nature07467

Keller, G.B., Mrsic-Flogel, T.D., 2018. Predictive Processing: A Canonical Cortical Computation. Neuron 100, 424–435. https://doi.org/10.1016/j.neuron.2018.10.003

Oliveira, A.S., Schlink, B.R., Hairston, W.D., König, P., Ferris, D.P., 2016. Proposing Metrics for Benchmarking Novel EEG Technologies Towards Real-World Measurements. Front. Hum. Neurosci. 10, 188. https://doi.org/10.3389/fnhum.2016.00188

O’Toole, S.M., Oyibo, H.K., Keller, G.B., 2023. Molecularly targetable cell types in mouse visual cortex have distinguishable prediction error responses. Neuron 111, 2918-2928.e8. https://doi.org/10.1016/j.neuron.2023.08.015

Rao, R.P.N., Ballard, D.H., 1999. Predictive coding in the visual cortex: a functional interpretation of some extra-classical receptive-field effects. Nat. Neurosci. 2, 79–87. https://doi.org/10.1038/4580

Vasilevskaya, A., Widmer, F.C., Keller, G.B., Jordan, R., 2023. Locomotion-induced gain of visual responses cannot explain visuomotor mismatch responses in layer 2/3 of primary visual cortex. Cell Rep. 42, 112096. https://doi.org/10.1016/j.celrep.2023.112096

Webb, J.M., Sohoglu, E., 2025. Cortical tracking of prediction error during perception of connected speech. https://doi.org/10.1101/2025.07.18.665498

West, C.L., Bastos, G., Duran, A., Nadeem, S., Ricci, D., Groves, A.M.R., Wargo, J.A., Peterka, D.S., Leeuwen, N.V., Hamm, J.P., 2024. A lasting impact of serotonergic psychedelics on visual processing and behavior. https://doi.org/10.1101/2024.07.03.601959

Version 2

Abstract

Likely the strongest predictor of visual feedback is self-motion. In mice, the coupling between movement and visual feedback is learned with first visual experience of the world (Attinger et al., 2017), and brief perturbations of the coupling result in strong visuomotor mismatch responses in visual cortex that possibly reflect prediction errors (Keller et al., 2012; Zmarz and Keller, 2016). In humans, predictive coding has primarily been studied using oddball paradigms which rely on violations of stimulus probability based on recent sensory history. It was still unclear, however, whether humans exhibit visuomotor mismatch responses similar to those observed in mice. This question was important for two reasons. First, visuomotor mismatch responses in humans constitute a basis to start translating the mechanistic understanding of the circuit that computes these responses from mouse to human cortex. Second, a paradigm that can trigger strong prediction error responses and consequently requires shorter recording times would simplify experiments in a clinical setting. Here, by combining a wireless EEG recording system with a virtual reality headset, we found robust visuomotor mismatch responses in human cortex that were characterized by a reversed polarity relative to visual evoked responses and a greater signal power than both visual responses and oddball mismatch responses.

Introduction

Predictive processing is a framework for understanding brain function. It proposes that the brain constructs internal models of the world based on regularities in past sensory input and sensorimotor loops to predict upcoming sensory input. The brain does this by minimizing the mismatches between predicted and actual sensory input. These mismatches, known as prediction errors, play a dual role in both updating internal representations and internal models (Friston, 2005; Keller and Sterzer, 2024; Rao and Ballard, 1999).

Predictions about sensory input can be made based on a variety of sources of other information the brain has available. Certain stimuli can be predicted from the preceding sensory input – experimentally this can be used, e.g., in oddball paradigms to trigger stimulus history based prediction errors (Garrido et al., 2009). Predictions can be crossmodal. Certain sounds are associated with specific types of visual input. Experimentally, visual stimuli can be coupled to sounds to then trigger audiovisual prediction errors in mouse visual cortex (Garner and Keller, 2022). Predictions can also be based on spatial memory. Not seeing a stimulus in a specific spatial location can result in visuospatial prediction errors in mouse cortex (Fiser et al., 2016). However, the strongest predictor - by frequency of experience, consistency in coupling, and precision in timing - is self-generated movement. Every movement is directly coupled to sensory feedback throughout life. Almost all visual and proprioceptive sensory input is the consequence of self-motion.

An essential ingredient to the computation of visuomotor prediction error responses are motor-related predictions of visual feedback (Keller and Mrsic-Flogel, 2018). In the mouse, evidence for such predictions has come from the discovery of a strong motor-related drive of activity in visual cortex (Keller et al., 2012; Niell and Stryker, 2010; Saleem et al., 2013). These motor-related signals are likely in part driven by motor-related predictions arriving from motor areas of cortex (Leinweber et al., 2017). However, the overall level of this motor-related activity is higher than one would expect simply from a motor-related prediction of visual input. Motor related activity is also evident in complete darkness and even prior to first visual experience in life (Attinger et al., 2017; Keller et al., 2012; Mahringer et al., 2022). It is possible, that this relatively high level of motor related activity in mouse visual cortex is a reflection of the fact that mice rely much less on fine visuomotor control.

In systems that have evolutionarily been driven to very high levels of precision of sensorimotor control, like vocal learning, there is much less such motor related activity. This is evident in the auditory system of a songbird, which relies on very temporally precise sensorimotor error detection for vocal learning, where there is much less motor related drive of auditory responses (Keller and Hahnloser, 2009). Similarly, there is very little movement related drive of activity in visual cortex of non-human primates (Liska et al., 2024; Talluri et al., 2023). In humans, movement-related modulation of EEG activity in visual cortex has been described (Cheng and Nordin, 2025; Gramann et al., 2010), but it is still unclear if visual responses are modulated by these signals. If indeed the evolutionary pressure for precision in visuomotor coupling determines the amount of motor modulation of visual responses, we would expect to find less of it in humans than in the mouse. Humans have a much higher reliance on visual feedback to guide movements than mice do.

This heightened reliance on visual guidance suggests that the human brain should be particularly adept at detecting mismatches in visuomotor coupling. Indeed, violations of the coupling between actual hand position and that of a virtual hand trigger responses in human visual cortex (Stanley and Miall, 2007). Even violations of simple action-outcome rules can trigger robust error responses in human subjects (Kimura and Takeda, 2019, 2014). Thus, we would also expect to find strong visuomotor mismatch responses in humans, comparable to – or stronger than – those observed in mice (Keller et al., 2012).

Results

Visual responses

To quantify visual responses in freely moving human participants, we combined a virtual reality (VR) headset with an 8-electrode wireless EEG recording system (Methods; Figure 1A). Participants were shown a virtual environment that consisted of a large empty floor space with a square checkerboard pattern fixed to their viewing angle directly in front of them (Figure 1B). The checkerboard covered a visual angle of 53° and contained a red fixation dot in the center. Participants were asked to fixate on the red dot throughout all measurements. The checkerboard reversed white-black at intervals sampled from a uniform distribution between 2 and 4 s (Figure 1C). Recordings were split into a passive session during which participants were seated on a chair, and an active session during which they were instructed to walk around a rectangular floor area of 5 by 7 m. During the entire experiment, we tracked the location of participants (Figure 1D).

EEG responses to visual stimulation are modulated by walking.

(A) Participant wearing the setup for wireless EEG recordings and a virtual reality headset. An Open BCI EEG electrode cap is combined with META QUEST 3. Inset: First person view of the virtual environment. Red fixation dot was enlarged in figure to make it visible. (B) A third person view of the virtual scene used to study visual responses to a reversing checkerboard pattern. The black line on the ground indicates the safety boundary within which the participant was instructed to walk around during the movement block of the paradigm. The grid indicates 1 by 1 m squares. (C) Participants viewed a reversing checkerboard in two sessions: During half of the trials, they remained seated, and during the other half they walked within a defined safety boundary. Checkerboard reversed colors at random intervals every 2 to 4 s. (D) 3D trajectory of an example participant during the visual experiment. Movement was recorded using accelerometers integrated into the VR headset. (E) Visually evoked potentials measured in the sitting session on the occipital electrodes O1 and O2. Responses recorded on other electrodes are presented in Figure S3. Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red line indicates the time of the checkerboard reversal. P1 - positive peak at around 88 ms. (F) As in E, but while participants were walking. Arrow indicates an early negative deflection preceding the P1 component. Inset: Comparison of visual evoked potentials measured in the sitting E and walking F sessions. The horizontal bar above the plot marks time bins in which responses differ significantly (black: p < 0.05) or do not differ (gray: p > 0.05).

Walking triggered movement related artifacts in the EEG signals (Figure S1). The strength of these movement artifacts varied as a function of hairstyle and gait of the participants. All participants were encouraged to go ‘gentle’ to reduce this problem. We excluded all data with eye blink and high movement related artifacts from further analysis (Figure S2, Methods; for number and percentage of excluded trials per session see Table S2).

We could detect multiphasic responses to the checkerboard inversion in occipital EEG electrodes (Figures 1E and S3). These responses were comparable to previous reports (Drislane, 2007) and exhibited a positive peak at around 88 ms (P1). Interestingly, when participants were walking, the same responses showed an early negative deflection with a peak latency of 48 ms that preceded P1 (Figure 1F). This early negative deflection at around 50 ms is also apparent in data from earlier studies comparing visual responses in walking and standing participants (Gramann et al., 2010).

While participants were instructed to fixate on the red dot, it is possible that there are systematic differences in eye movements between walking and sitting sessions. Our setup did not allow for concurrent eye tracking; thus, we cannot exclude the possibility that differences in eye movements contribute to the differences in EEG responses we observe. However, given the relatively rapid onset of response differences (40 ms), we suspect that stimulus-triggered eye movements cannot account for these early differences.

Visuomotor mismatch responses

To measure visuomotor mismatch responses, participants were instructed to walk around in a virtual corridor (Figure 2A). The corridor was 1 m wide, 2.4 m high and of an oval shape covering approximately 7 by 5 m (Figure 2B). Virtual movement in the corridor was coupled to the movement of participants. We refer to this as closed loop coupling. To trigger visuomotor mismatches, we briefly (0.5 s) halted the coupling between movement of the participants and visual feedback in the virtual corridor at random times (every 10 to 15 s; Methods; Figure 2C). Throughout all experiments, we tracked the virtual position of participants as they were walking around the corridor (Figure 2D).

Visuomotor mismatch elicits strong EEG responses.

(A) Participant navigating a virtual tunnel. Inset: First person view of the tunnel. (B) Third person view of the virtual environment used to study visuomotor mismatch responses. The grid indicates 1 by 1 m squares. (C) In closed loop sessions, the walking speed of participants was coupled to movement in the virtual corridor. Visuomotor mismatches were introduced by briefly halting the visual flow for 0.5 s. Following each visuomotor mismatch event, the view was updated to the participants’ current position (brief peaks in visual flow speed after the mismatch event) and visuomotor coupling was resumed. (D) Trajectory of an example participant during the visuomotor mismatch paradigm. (E) Responses to visuomotor mismatches recorded from occipital electrodes (O1 and O2). Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset and offset of the visuomotor mismatch. The gray shaded areas mark the analysis windows used to quantify response strength in G and H. Analysis windows of 100 ms were centered on the peak of the visuomotor mismatch response and the teleport response. (F) As in E, but for visual flow playback halt responses recorded from occipital electrodes. (G) Comparison of the response strength to visuomotor mismatches and playback halts. Here and elsewhere: Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from one participant. ***: p<0.001. See Table S1 for all statistical information. (H) Comparison of the response strength to teleport events following visuomotor mismatches in the closed loop session and playback halts in the open loop session. n.s.: not significant. See Table S1 for all statistical information. (H) Average walking speed of participants during visuomotor mismatches. Inset: Temporally expanded view. The horizontal bar above the plot marks time bins where walking speed differs significantly (black: p < 0.05) or does not differ (gray: p > 0.05) from baseline.

Anecdotally, the perception some participants reported experiencing during these visuomotor mismatches was a very salient sudden movement of the entire visual world. In the words of one participant: “The world suddenly flew forward!”. This was reminiscent of a case report of a patient with a lesion to the lateral rectus muscle described by von Helmholtz (von Helmholtz, 1867): The lateral rectus muscle moves the eye laterally. The patient, upon attempting to move the affected eye laterally, reported seeing the world rapidly moving in the direction of the intended eye movement. One interpretation of this is that the combination of movement and absence of resulting visual flow change results in the perception that the world must be moving.

In the EEG recordings of occipital electrodes, we found strong responses triggered by these visuomotor mismatches (Figure 2E). Responses were dominated by a positive component peaking at 180 ms. Following the 0.5 s halting of the visual stimulus, the virtual location was updated to match the actual location of participants (we call this a teleport event), and the visual flow coupling was resumed. This resulted in a sudden change of visual stimulus combined with a visual flow onset and drove what is likely a visual response in the EEG signal (Figure 2E).

To quantify how much of the visuomotor mismatch response can be explained by visual input alone, we exposed participants to a replay of the visual flow that was self-generated in the preceding closed loop session, including the visual flow halts that constitute visuomotor mismatches in the closed loop session. We refer to this as open loop coupling. For these experiments, participants were seated on a chair. To reduce the likelihood of triggering nausea in participants, roll and pitch movements of the head were removed from this replay. We found measurable responses to visual flow playback halts (Figure 2F), but these were much smaller than visuomotor mismatch responses (Figure 2G). In the few participants who volunteered to experience playback that included pitch and roll movements, playback halt responses were also significantly smaller than visuomotor mismatch responses (Figure S4A-S4C) and not different from those in the reduced open loop session (Figure S4D and S4E). The difference in response amplitude between mismatch and playback response could not be explained by a simple movement related gain of visual responses, as the response to the teleport and re-onset of visual flow was similar in both closed and open loop sessions (Figure 2H). Note that the mismatch responses and playback halt responses shown here do not come from fully overlapping pools of participants (Table S1). A subset of visuomotor mismatch sessions had to be excluded because of recording noise, while some participants aborted the open loop recording due to developing nausea (Methods; for number and percentage of excluded trials per session see Table S2). A participant-wise comparison of mismatch and playback halt responses is shown in Figure S5.

By design, open loop sessions in the form of an identical playback (to match the visual statistics of open and closed loop sessions) always followed closed loop sessions, which could mean that the difference between visuomotor mismatch and playback halt responses could be driven by some form of adaptation. To control for this, we repeated the experiment in a separate cohort of participants with the order of sessions reversed: Participants first viewed a playback of the closed loop session of the previous participant and only then performed an active closed loop session. We found that independent of the order of closed and open loop sessions, visuomotor mismatch responses were larger than playback halt responses (Figure S6), thus excluding the possibility that the response difference may be driven by adaption or a reduction in salience of the sudden flow halt.

Although participants reduced their walking speed in response to the visuomotor mismatch when averaged over all trials, this did not reach statistical significance (Figure 2I). But see Figure 4A, for evidence that there was a significant reduction in walking speed when analyzing only the first few mismatch trials. In either case however, the reduction in walking speed was of much longer latency compared to the visuomotor mismatch responses, demonstrating that walking speed changes cannot explain visuomotor mismatch responses.

Finally, to test whether visuomotor mismatch events trigger eye movements that could explain visuomotor mismatch responses, we measured eye movements triggered by visuomotor mismatch in an independent cohort of participants. As with the EEG experiments, participants were exposed to a closed loop session with visuomotor mismatches. Throughout the experiment, we tracked eye position and eye blinks using an eye tracker built into the VR headset (Methods). Note, this VR headset was different from the one we used in the EEG experiments. We found that visuomotor mismatch triggered detectable eye responses. While we could not detect any systematic eye movements triggered by mismatch (Figure S7A and S7B), participants exhibited a reduction in average eye velocity (Figure S7C) and a reduction in eye blink rate (Figure S7D) with a latency of about 208 ms and 541 ms, respectively. Thus, both of these changes had latencies longer than those we observed for responses to visuomotor mismatch in the EEG experiments (Figure S7E). Thus, we find no evidence that visuomotor mismatch responses can be explained by eye blinks or eye movements.

Distribution of visuomotor mismatch response strength

In mice, visuomotor mismatch responses originate in primary visual cortex and propagate to a network of areas across dorsal cortex (Heindorf and Keller, 2023; Takeuchi et al., 2024). Thus, we expected to find larger and faster responses in the two occipital electrodes. To test this, we compared the visuomotor mismatch responses across the eight recorded locations in both response strength and timing (Figures 3A and 3B). We indeed found that the strongest responses were observed at occipital electrodes O1 and O2, though significant responses were also present at frontal (Fp1-2) and parietal (P3-4) sites (Figure 3C). We found a significant difference in response latency between occipital and frontal electrodes, with occipital signals occurring significantly earlier than frontal responses. (Figures 3D and 3E). This suggests that mismatch processing may initially arise in sensory visual areas before engaging higher-order frontal regions.

Visuomotor mismatch responses are most prominent over occipital cortex.

(A) Top down view of EEG electrode locations on the head. (B) Visuomotor mismatch responses measured on electrodes shown in A. Solid black lines represent the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (C) Comparison of the response strength to visuomotor mismatches measured on electrodes shown in A. Average response strength was calculated within a 100 ms window centered on the peak of the average visuomotor mismatch response across all electrodes. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from an individual participant. ***: p<0.001, **: p<0.01, *: p<0.05, n.s.: not significant. See Table S1 for all statistical information. (D) The responses shown in B, averaged over pairs of electrodes and overlaid. Solid black lines represent mean and shading SEM across electrodes. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (E) Comparison of the latency to half maximum response for the 4 electrode pairs. **: p<0.01, n.s.: not significant. See Table S1 for all statistical information.

Visuomotor mismatch responses show experience-related changes predominantly over frontal electrodes.

(A) Average walking speed during visuomotor mismatch events in early versus late trials. The solid black line shows the mean and the shaded area the SEM across participants. Vertical red dashed lines indicate mismatch onset and offset. The horizontal bar above the plot marks time bins in which the walking speed in early trials differs significantly (black, p < 0.05) or not (gray, p > 0.05) from that in late trials. (B) Visuomotor mismatch responses in early versus late trials over frontal electrodes. (C) As in B, but for occipital electrodes. (D) Comparison of the difference between visuomotor mismatch response in early and late trials for frontal and occipital electrodes. Average response strength was calculated within a 100 ms window centered on the peak of the average visuomotor mismatch response in early trials above frontal or occipital electrodes. Boxes mark median, quartiles, and range of data not considered outliers. *: p<0.05. See Table S1 for all statistical information.

Temporal dynamics of neuronal and behavioral responses to visuomotor mismatch across a session

To examine how mismatch responses evolved over the session, we compared responses in early (first five) and late (last five) trials. We found that participants reduced their walking speed with a delay of about 500 ms following mismatch events. This effect was more prominent in early trials (14.3%) compared to late trials (5.7%) (Figure 4A). Interestingly, visuomotor mismatch responses mirrored this pattern over frontal electrodes with a clear attenuation from early to late trials (Figure 4B). By contrast, response changes in occipital electrodes were smaller between early and late trials (Figure 4C). The fact that the reduction over frontal cortex is larger than that of responses over visual cortex (Figure 4D), indicates that there may be a gradient of experience dependent changes in processing from early sensory to frontal areas of cortex that relate to changes in behavioral responses to, but not detection of prediction errors.

Comparison of visual responses and visuomotor mismatch responses

The time course of visual responses (Figure 1E) and visuomotor mismatch responses (Figure 2E) differed in that they appeared to exhibit reversed polarity (Figure 5A). Despite this difference in polarity, the initial peaks occurred at similar latencies. The total power of the visuomotor response was higher than that of the visual response (Figure 5B), but given that we did not optimize either the visuomotor mismatch stimulus nor the visual stimulus for maximum response, this comparison is more qualitative than quantitative in nature. We include it here primarily to illustrate how the visuomotor mismatch response compares to a more conventional visual response.

Visuomotor mismatch responses have reversed polarity and more power compared to visual responses.

(A) Comparison of visual (Figure 1E) and visuomotor mismatch (Figure 2E) responses recorded from all electrodes. Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset (visual and mismatch) and offset (mismatch) of the stimuli. (B) Comparison of the power of visual and visuomotor mismatch responses, calculated within a 0 - 0.5 s time window following stimulus onset. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from an individual participant. ***: p<0.001, *: p<0.05. See Table S1 for all statistical information.

Comparison of visuomotor mismatch responses and auditory oddball mismatch responses

To further contrast the strength of visuomotor mismatch to other frequently used prediction error signals, we also recorded EEG responses in an auditory oddball paradigm. Mismatch responses recorded with this paradigm can be thought of as a stimulus history prediction error (Garrido et al., 2009; Näätänen et al., 2007). We implemented an oddball paradigm composed of a sequence of identical tones that was occasionally disrupted by a tone of a different frequency (Figure 6A). We used pure tones of 1 kHz and 1.2 kHz frequency and alternated in blocks which tone was standard and which was the deviant (Methods). Participants watched short silent movies in the VR headset while listening to the tone sequences delivered in headphones. To isolate the oddball driven mismatch response, we subtracted the response to the tone when presented as a standard from the response to the same tone when presented as a deviant. The resulting difference waveform closely resembled those reported in previous studies (Figures 6B-6C, S8, and S9) (Näätänen et al., 2007). Comparing these oddball mismatch responses to the visuomotor mismatch response in occipital electrodes, we found that both had similar dynamics with a dominant positive peak at around 180 ms (Figure 6D). Comparing the total power of the two responses, we found that visuomotor mismatch responses were significantly larger than oddball mismatch responses (Figure 6E). This was also true for responses measured in temporal electrodes (Figures S8 and S9). But also here, the comparison of total power is more qualitative than quantitative as neither stimulus was optimized for maximum EEG response, and we include it to illustrate how typical variants of these responses compare.

Visuomotor mismatch responses are larger than auditory oddball mismatch responses but have similar temporal dynamics

(A) Design of the auditory oddball paradigm with examples of the silent films participants were exposed to. (B) Top: Auditory responses to the 1 kHz tone presented as a standard versus as a deviant. Bottom: Oddball mismatch response calculated by subtracting the average response over trials in which the tone was presented as a standard from those when the tone was presented as a deviant. Solid black lines represent the mean, and shading indicates the SEM across participants. Dashed vertical red line is the onset of the auditory stimulus. (C) As in B, but for the 1.2 kHz tone. (D) Comparison of visuomotor mismatch, oddball mismatch (average over data shown in panels B and C) and playback halt responses recorded from occipital electrodes. Solid black lines represent the mean response, with shading indicating SEM across electrodes. Dashed vertical red lines are onset (visual, mismatch) and offset (mismatch) of the stimuli. (E) Comparison of the power of visuomotor mismatch, oddball mismatch response (average over data shown in panels B and C) and playback halt responses, calculated within a 0 s - 0.5 s time window following stimulus onset. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from an individual participant. ***: p<0.001; *p<0.05. See Table S1 for all statistical information.

Discussion

The utility of paradigms for measuring prediction errors in humans primarily comes from their relevance to psychiatric disorders, where disruptions in prediction error processing are widely implicated (Adams et al., 2013; Fletcher and Frith, 2009; Kirihara et al., 2020; Qela et al., 2025; Sterzer et al., 2018). Various types of prediction error responses are being explored as potential biomarkers in clinical research, serving to monitor disease progression, evaluate symptoms, and provide insight into the underlying neuronal mechanisms of these conditions.

The most prominent of these prediction errors is mismatch negativity (MMN). MMN is the first component of the oddball mismatch response (Näätänen et al., 1978). MMN has been shown to be reduced or abnormal in patients with schizophrenia spectrum disorders (Todd et al., 2012; Umbricht and Krljes, 2005), autism spectrum disorders (Chen et al., 2020; Dunn et al., 2008; Schwartz et al., 2018), speech disorders (El Hatal de Souza et al., 2020), depression (Tseng et al., 2021), and bipolar disorders (Chitty et al., 2013). More generally, it is thought that many of these disorders can be related to changes in predictive processing. In the case of schizophrenia, for example, the symptoms are thought to arise from an imbalance in the strength of high and low level predictions (priors) (Schmack et al., 2017, 2015a, 2015b, 2013). Further testing these ideas will require experiments that trigger identified functional responses in humans, for which we have a circuit level understanding based on cell type specific recordings.

We were able to measure robust visuomotor mismatch responses in freely moving humans (Figure 2). These signals cannot be attributed to changes in participants’ behavior or to simple visual offset responses. Notably, the visuomotor mismatch responses exhibited a markedly different temporal profile compared to purely visual responses (Figure 5). What might account for this difference? One possibility is that visuomotor mismatch signals may rely more strongly on input from other cortical areas than visual responses and are thus delayed relative to visual responses. A more interesting interpretation is that visual responses and visuomotor mismatch responses are both prediction errors of different types. A central tenet of the cortical circuit for predictive processing is the split into separate populations of neurons that compute positive and negative prediction errors (Keller and Mrsic-Flogel, 2018; Rao and Ballard, 1999). In this interpretation, a visuomotor mismatch response is a negative prediction error, while the response to a visual stimulus is a positive prediction error.

Visuomotor prediction errors are likely computed in layer 2/3 of mouse visual cortex (Jordan and Keller, 2020), and are also preferentially detectable in superficial layers of human visual cortex (Thomas et al., 2024). In the mouse, positive and negative prediction errors are likely computed in separate populations of neurons in visual cortex (Keller and Mrsic-Flogel, 2018; O’Toole et al., 2023). These two cell types have a different depth distribution, with negative prediction error neurons more superficial and positive prediction error neurons located deeper in cortex (O’Toole et al., 2023). Given that EEG signals are thought to exhibit positive or negative deflections as a function of depth of the source (Cohen, 2017; Kirschstein and Köhling, 2009), the polarity reversal of visual and visuomotor mismatch responses (Figure 5) may be the consequence of different populations of cells being activated and inhibited. Visuomotor mismatch should activate negative prediction error neurons and inhibit positive prediction error neurons, while the reverse is the case for a visual stimulus.

The fact that there is also a strong visual response to the visual flow re-onset following visuomotor mismatch means that our visuomotor mismatch paradigm might allow us to measure both negative and positive prediction error responses. More intriguingly, there is a differential modulation by walking of the two responses (Figures 2G and 2H). In theory, visuomotor mismatch drives a combination of two negative prediction errors. One is based on a movement related prediction. The other on a stimulus-history related prediction. Both the self-motion as well as the ongoing visual flow would predict a continuation of visual flow. A visuomotor mismatch violates both predictions. During passive observation, there is no movement-related prediction, but the stimulus history-based prediction is still violated.

Intriguing is the similarity in timing and polarity of the playback halt and oddball response (Figure 6D). An oddball response is always a combination of both a positive and a negative prediction error. There is a negative prediction error for the absence of the expected standard tone as well as a positive prediction error for the presence of the unexpected oddball tone. Thus, it is conceivable that the subtraction of standard response from the oddball response, isolates the component of the oddball response that corresponds to a negative prediction error.

One of the key challenges in systems neuroscience is translating findings from animal models to humans. Although recent animal studies have provided detailed insights into the circuit level implementation of predictive processing in the cortex (Keller and Mrsic-Flogel, 2018), and psychological and psychiatric conditions have long been described within this framework (Keller and Sterzer, 2024; Sterzer et al., 2018), translation to human research has been limited by a lack of methods to record cell type specific signals in human experiments. With an understanding from animal models of how positive and negative prediction errors are computed, and which cortical layers are involved, an approach based on functionally identified responses might be most promising. One step in this direction is to use experimental paradigms that can cleanly separate e.g. positive and negative prediction error responses. As argued above, we speculate that the visuomotor mismatch paradigm as we have used it here, is one possible way to achieve this.

It should also be highlighted that the paradigm we have developed here is still limited in its clinical applicability by practical constraints, including movement-related artifacts, signal-to-noise levels, and the requirement for active engagement. Nevertheless, active task engagement may itself be a critical feature for revealing deficits in cortical processing that remain undetectable in more passive paradigms. Thus, although active, more video-game-like paradigms may appear less practical in certain clinical contexts, they may be essential for capturing meaningful and behaviorally relevant neuronal responses. Thus, methodological refinement and hardware optimization will be required before reliable clinical translation. However, these limitations do not detract from the paradigm’s immediate utility as a research tool for studying prediction error processing.

Methods

Participants

The study was approved by Ethikkommission Nordwestund Zentralschweiz (Project-ID: 2024-02458). Participants signed a written informed consent before participation and were not financially compensated for their participation. A total of 91 recording sessions from healthy adults were included in the study. Sessions were obtained from participants aged 18–65 years, with a small number of participants contributing more than one recording session. The age distribution across sessions was as follows: 38 sessions from ages 18–30, 37 sessions from ages 31–40, 11 sessions from ages 41–50, and 5 sessions from ages 51–65. None of the participants reported a prior diagnosis of movement disorders, vestibular dysfunction, or epilepsy. Experience with virtual reality technology ranged from beginner to advanced, with most participants reporting minimal prior experience.

EEG recordings in humans

We integrated a wireless EEG recording system with a VR headset. The EEG recording system was composed of a wet electrode cap and a Cyton biosensing board from OpenBCI. This allowed us to record 8 EEG channels (43 sessions: Fp1, Fp2, C3, C4, P3, P4, O1 and O2; 48 sessions: Fp1, Fp2, T3, T4, T5, T6, O1 and O2) at a 250 Hz sampling rate. Electrode labels follow the international 10-20 system: Fp = frontopolar, C = central, P = parietal, O = occipital, and T = temporal; odd numbers indicate the left hemisphere, and even numbers the right (Figure 3A). All EEG data were wirelessly (via Bluetooth) transmitted to a nearby computer. The VR headset was a Meta Quest 3, with a virtual environment developed in the Unity engine (Unity Technologies). Virtual 3D objects were designed in Fusion 360 (Autodesk). Synchronization between EEG recording and VR headset was performed by connecting an auditory output of the VR headset to the Cyton board to exchange synchronization triggers. This allowed us to synchronize EEG data with VR events offline. Auxiliary signals, including the participant’s position, stimulus trigger timing, and type, were recorded directly on the headset at a sampling rate of approximately 100 Hz.

Eye movement measurements

We recorded eye blinks and eye movements using a Meta Quest Pro VR headset and its built in eye tracker during closed loop sessions. Eye-tracking data were sampled at 72 Hz. We combined blink signals from left and right eye, by averaging the two channels at each time point, yielding a single blink/eye-closure trace per participant. To compute eye movement speed, we first computed eye-movement speed separately for the left and right eye from the moment-to-moment change in eye rotation between consecutive samples and then averaged the two speed traces to obtain a single eye-speed measure.

Visual responses

Visual stimuli were presented using the VR headset. We used a reversing square checkerboard stimulus to drive visually evoked potentials. The checkerboard reversed colors at random intervals (between 2 and 4 s). In virtual space, the checkerboard measured 0.5 by 0.5 m and was positioned 0.5 m in front of participants’ eyes. This resulted in a horizontal and vertical coverage by the checkerboard of approximately 53° of visual angle. Each visual stimulation session lasted 4 min, during which the checkerboard reversed colors between 78 and 82 times. For half of the session, participants viewed the stimulus while seated; for the other half, they were instructed to walk freely within a 7 by 5 m empty floor space. The order of sitting and walking sessions was randomized across participants.

Visuomotor mismatch responses

For the experiments measuring visuomotor mismatch responses, we used a 3D virtual corridor with vertical gratings on the walls. The corridor measured 1 m in width, 2.4 m in height, and had an oval shape of 7 by 5 m. Visuomotor mismatches were introduced at random intervals every 10 to 15 s as participants walked through the corridor. The session lasted 5 min and included between 22 and 26 visuomotor mismatch events. During these events, the coupling between the participant’s movement and the visual feedback in the VR headset was briefly interrupted, i.e. the visual scene was frozen for 0.5 s. Because participants continued to move during this time, they were teleported to their current position in the virtual space following visuomotor mismatch event.

Playback halt responses

To quantify how much of the visuomotor mismatch response could be explained by visual input alone, participants were asked to passively observe a replay of the visual flow they had self-generated during the preceding closed loop session; we refer to this replay condition as the open loop session. We quickly learned, however, that watching 5 minutes of playback in the VR headset triggered nausea in most participants. Thus, we started experimenting with changes to the playback to minimize the risk of triggering nausea. One such modification we settled on to use for experiments was a playback version that omitted pitch and roll movements of the head. Full playback involved six degrees of freedom (6DOF): 3D position in space, plus pitch, yaw, and roll angles of the head. The constrained playback included only four degrees of freedom (4DOF): The 3D position in space, but only yaw movements of the head. A subset of participants volunteered to view the full playback (Figure S4). Open loop sessions were conducted while participants were seated and lasted 5 min, matching the duration of the closed loop session. Approximately 23% of participants reported strong nausea during the beginning of the 4DOF open loop session, at which point the recordings were terminated.

Mismatch negativity

To measure mismatch negativity, we used an auditory oddball paradigm. Auditory stimuli consisted of two pure tones: 1 kHz and 1.2 kHz. They were presented at 60 to 70 dB sound pressure level in a counterbalanced design, in which each tone served as the standard in one condition and as the deviant in the other. Tones were 50 ms in duration, including 5 ms linear rise and fall ramps. Stimuli were delivered in a pseudorandom order, with deviant tones comprising 10% of all trials and preceded by at least five standard tones. The inter-trial interval was selected from a uniform distribution of between 500 and 600 ms. Each session lasted 3 min and included between 28 - 33 deviant and 244 - 299 standard stimulus presentations. Participants were seated and listened to the tones while watching a silent movie that was not related to the tone sequences.

EEG Signal Analysis

Data analysis was done using custom-written MATLAB scripts. EEG signals were band-pass filtered between 1 and 100 Hz. To remove power line noise, a band-stop filter was applied between 40 and 60 Hz. Movement of the participants triggered all varieties of movement related artifacts in the EEG recordings. The strength of these artifacts depended on a variety of factors: Impedance of the electrodes, hair style of participants, gait pattern, and likely others. To reduce data contaminated by excessive movement artifacts and eye blinks, trials with a maximum absolute response amplitude exceeding 100 µV within the time window −0.5 to 1 s relative to the trigger were discarded from further analysis (Figure S2). Data from each electrode were included in the final analysis only if at least 15 triggers remained after exclusion of triggers with excessive movement artifacts (Table S1). To compare average response strength across sessions, a 100 ms analysis window was used, centered on the peak of the respective responses: The visuomotor mismatch event recorded at occipital electrodes O1 and O2 (Figures 2G, S4C, S5C, S6C), the visual flow re-onset response (Figure 2H), the visuomotor mismatch response in early trials for frontal and occipital electrodes (Figure 4D), the mean visuomotor mismatch response across all electrodes (Figure 3C), the mean visual response across all electrodes (Figure S3C) or playback halt response in the 6DOF session (Figure S4E). Signal power was compared by calculating the mean squared amplitude within a 0 - 0.5 s analysis window following stimulus onset (Figures 5B, 6E, S8E, S9D). To analyze how responses evolved over the course of the session, we computed moving-average responses across 10 trials at successive percentages of progress through the paradigm (Figure S6D and S6E).

Statistical tests

All statistical analyses were conducted using hierarchical bootstrap (Saravanan et al., 2020). Bootstrap resampling enables statistical comparisons across sessions without assuming a specific distribution of the EEG data. For analysis in Figures 1, 2, 4, 6, S4, S5 and S6, we averaged signals across electrode pairs (O1-2 or Fp1-2) and treated the result as a single data point per participant. For Figures S8 and S9, the same procedure was applied to temporal electrode pairs (T3-4 and T5-6). Likewise, in Figures 3D, 3E, and 5, we averaged signals from the respective electrode pairs O1-2, C3-4, P3-4, and Fp1-2, and treated each pair as a single data point per participant. For the analysis shown in Figures 3C and S3C we included signals from all electrodes and used a nested bootstrap to account for multiple data points originating from the same participant. We first resampled the data with replacement at the level of participants, followed by resampling at the level of electrodes. For each resampled population, we computed the mean response and repeated this procedure 10000 times. The p value was estimated as the fraction of bootstrap samples in which the sample mean violated the tested hypothesis. See Table S1 for all information on number of participants or electrodes used for all analyses shown.

eLife Assessment

This valuable study demonstrates that self-motion strongly affects neural responses to visual stimuli, comparing humans moving through a virtual environment to passive viewing. The evidence for visuomotor mismatch responses is solid, although the interpretation in terms of prediction remains somewhat preliminary. This study bridges human and rodent studies on the role of prediction in sensory processing, and is therefore expected to be of interest to a large community of neuroscientists.

Reviewer #1 (Public review):

In this paper, Solyga, Zelechowski & Keller study human visuomotor mismatch responses as an alternative instantiation of prediction errors to classic oddball paradigms. Using VR, they created a condition in which participants were moving around thereby creating a visuomotor coupling between physical movement and visual flow. To attempt to isolate the contribution of specifically movement-related predictions in this condition, they contrasted it to a condition in which participants were seated and rewatching their movement trajectory during the 'active' condition. Visuomotor mismatches were created by temporarily decoupling movement and visual experience by halting the VR display as participants continued to move.

The core finding of the paper is that participants exhibit a positively-valenced response to the visuomotor decoupling in the active but not in the passive condition. Since walking speed only insignificantly slows down following decoupling events in the active conditions, the authors argue that this difference can not be accounted for by "changes in participants' behavior or to simple visual offset responses" with the latter being equal across both conditions. The following reinstatement of the coupling in turn does not differ between the two conditions. The authors additionally show that this mismatch response differs from visual onset responses elicited by checkerboard inversions and that it's "qualitatively" stronger than more commonly studied auditory oddball mismatch responses.

The design with its focus on ecological validity is impressive, well-rationalized and the results are well illustrated. I additionally appreciate the control analyses with regards to changes in walking speed and playback DOF and, now added, additional participants who experience the passive condition before the active. I have a couple of questions/comments.

My main question in round 1 regarded the isolation of visuomotor mismatch. Although the comparison with a seated control seems like a very sensible way to control for simple visual responses, there seem to be more differences than just a break in visuomotor coupling between the conditions. I therefore wonder whether the reduced offset response in the seated condition may be, in part, explained differently. For example, given that participants always conduct the active condition before rewatching their movement in the seated condition, it seemed likely that there is a component of learning across the session that flow will sometimes be halted. This is confirmed with the analyses. The explanation that there is a visuomotor component here is given further weight by their conduction of an additional group of participants who perform the conditions in the reverse order, so this has strengthened the manuscript considerably. However, it does of course remain an imperfect control because the visual stimulus is now different between the conditions for these participants. It's the best that can be achieved with this type of paradigm though and of course it yields a great deal of ecological validity.

I was also wondering whether the authors may consider the findings in frontal electrodes more closely given that the title of the paper focuses on a specifically occipital effect. Their further analyses have confirmed that there are likely interesting frontal effects. From a theoretical point of view, the spatial dissociation in adaptation effects, which were stronger in frontal and weaker in occipital areas, seems interesting and perhaps worth discussing, especially given the interpretation that "mismatch processing may initially arise in sensory visual areas before engaging higher-order frontal regions." How come the frontal decrease in responses is not accompanied by an analogous decrease in its supposed occipital source? Could these two responses reflect different kinds of prediction error signals (i.e. objective vs subjective)?

I remain concerned that the authors fight too defensively that they have absolutely isolated visuomotor prediction mechanisms with this paradigm. It's a nice, informative study, but it seems odd to argue there are no other possible explanations. One picks a design to optimize some features but they will always come at some cost to others. Prioritising ecological validity, which is a justifiable aim, necessarily usually weakens some control over confounds.

To outline my reasoning fully: My concerns wrt generic influences of action on perception are reflected in Fig 1. The P1 is smaller when walking than sitting. It seems likely that the mismatch response reflects something about extrapolation or prediction, because it is larger when walking. However, it's not necessarily sensorimotor prediction. Even if you remove action from the equation, the flow can be extrapolated or predicted most of the time in a way it cannot so well when the video is halted. Of course the sitting condition somewhat controls for it, but when it came second the visual flow disruptions were more predictable here. A reduction in effects over time is indeed confirmed with their analyses. They now have conducted a study with the conditions in the reverse order and they find the same thing. But of course this necessitates non-identical visual flow because the sitting condition is playing the previous participant's flow. So it is likely that across all of these comparisons, it is the visuomotor mismatch that is especially salient. It's just that each comparison is a bit messy/confounded. It would strengthen the manuscript if there were some consideration given to the other processes likely at play here.

As a more minor point in response to our previous review, whether particular accounts represent an 'orthodox' view at present does not determine whether they raise logical issues in need of consideration. The authors may have missed that the papers in question consider mechanisms underlying the attenuation of particular pieces of information *from perception*. Not perceptual processing. We have one percept at any one moment in time and must understand how different population types synergistically generate that percept.

Similarly a little strange is the way in which the authors aggressively defend the position that self-generated motion is 'the strongest' type of prediction. Sure, we probably experience the effects of our actions more often than ambulances. But what about objects obeying laws of gravity or others' faces being structured and moving in systematic ways? It is hard to quantify, such that presumably many scientists would be skeptical of such a claim, and it is not needed logically to justify the importance of examining mechanisms enabling action to shape perceptual processing. I'd assume it better to fight the battles you need to (and can) fight, such that the robust claims carry more weight.

Hope these comments are helpful.

Reviewer #2 (Public review):

Summary:

This study investigates whether visuomotor mismatch responses can be detected in humans. By adapting paradigms from rodent studies, the authors report EEG evidence of mismatch responses during visuomotor conditions and compare them to visual-only stimulation and mismatch responses in other modalities.

Strengths:

- Authors use a creative experimental design to elicit visuomotor mismatch responses in humans.

- The study provides an initial dataset and analytical framework that could support future research on human visuomotor prediction errors.

Weaknesses:

- Methodological issues (e.g., volume conduction) make it difficult to confidently attribute the observed mismatch responses to activity in visual cortical regions. This could be alleviated by increasing the number of channels.

The authors successfully demonstrate that visuomotor mismatch paradigms can, in principle, be applied in human EEG. This approach provides a translational bridge between rodent and human work on predictive processing.

Reviewer #3 (Public review):

Solyga, Zelechowski, and Keller present a concise report of an innovative study demonstrating clear visuomotor mismatch responses in ambulating humans, using a mobile EEG setup and virtual reality. Human subjects walked around a virtual corridor while EEGs were recorded. Occasionally, motion and visual flow were uncoupled, and this evoked a mismatch response that was strongest in occipitally placed electrodes and had a considerable signal to noise ratio. It was robust across participants and could not be explained by the visual stimulus alone.

This is an important extension of their prior work in mice, and represents an elegant translation of those previous findings to humans, where future work can inform theories of e.g. psychiatric diseases that are believed to involve disordered predictive processing. For the most part, the authors are appropriately circumspect in their interpretations and discussions of the implications. The paper in its current form represents an important addition to the literature.

The authors have included analyses of the auditory mismatch using temporal electrodes, referenced to Cz (and therefore should exhibit a mismatch positivity). This added data clearly and convincingly shows that the sensorimotor mismatch is, indeed, stronger than the passive auditory MMN.

- The reference electrode placed at Cz makes it is difficult to interpret relative differences between frontal and occipital electrode responses, as the occipital electrodes are placed farther away from the Cz reference than the frontal electrodes. Similarly, signal occuring cortically near the Cz reference might only appear as though it is occipitally distributed in this montage. It is common in EEG research to re-montage the data to an averaged common reference in order to better interpret the scalp distributions. As the electrode coverage was sparse for some subjects, this could be challenging, and this reviewer does not feel that it is necessary to do this analysis step, or even to drastically rewrite the body of the paper. We only request that some discussion, however brief, is included in the discussion section or the methods that recommend more dense electrode coverage in the future to better interpret scalp distributions and potential meso-scale sources.

- This is just a suggestion. The authors are encouraged to analyse (and report) time-frequency power and phase locking for these mismatch responses, as is common in much of the literature (see Roach et al 2008 Schizophrenia Bulletin). This is not to say that doing so will yield insights into oscillations per se, but converting the data to the time-frequency domain provides another perspective that has some advantages. fosters translations to rodent models, as ERP peaks do not map well between species, but e.g. delta-theta power does (see Lee et al 2018 Neuropsychopharmacology; Javitt et all 2018 Schizophrenia research; Gallimore et al 2023 Cereb Ctx). Further, ERP peaks can be influenced by the actual neuroanatomy of an individual (especially for quantifying V1 responses). Time frequency analyses may aid in interpreting the "early negative deflection with a peak latency of 48 ms " finding as well. As it stands, the report is complete, and it would be acceptable if the authors chose to save this type of analysis for a future publication.

Author response:

The following is the authors’ response to the previous reviews

We thank you for the time you took to review our work and for your feedback! The main changes to the manuscript are:

We added a paragraph to the Discussion addressing differences in visuomotor mismatch responses recorded over frontal and occipital electrodes, and their possible interpretation.

We added time-frequency power and phase-locking analysis as supplementary figures to the manuscript.

We added a statement in the Discussion emphasizing the importance of performing these experiments with denser EEG channel coverage.

Public Reviews:

Reviewer #1 (Public review):

In this paper, Solyga, Zelechowski & Keller study human visuomotor mismatch responses as an alternative instantiation of prediction errors to classic oddball paradigms. Using VR, they created a condition in which participants were moving around thereby creating a visuomotor coupling between physical movement and visual flow. To attempt to isolate the contribution of specifically movement-related predictions in this condition, they contrasted it to a condition in which participants were seated and rewatching their movement trajectory during the 'active' condition. Visuomotor mismatches were created by temporarily decoupling movement and visual experience by halting the VR display as participants continued to move.

The core finding of the paper is that participants exhibit a positively-valenced response to the visuomotor decoupling in the active but not in the passive condition. Since walking speed only insignificantly slows down following decoupling events in the active conditions, the authors argue that this difference can not be accounted for by "changes in participants' behavior or to simple visual offset responses" with the latter being equal across both conditions. The following reinstatement of the coupling in turn does not differ between the two conditions. The authors additionally show that this mismatch response differs from visual onset responses elicited by checkerboard inversions and that it's "qualitatively" stronger than more commonly studied auditory oddball mismatch responses.

The design with its focus on ecological validity is impressive, well-rationalized and the results are well illustrated. I additionally appreciate the control analyses with regards to changes in walking speed and playback DOF and, now added, additional participants who experience the passive condition before the active. I have a couple of questions/comments.

My main question in round 1 regarded the isolation of visuomotor mismatch. Although the comparison with a seated control seems like a very sensible way to control for simple visual responses, there seem to be more differences than just a break in visuomotor coupling between the conditions. I therefore wonder whether the reduced offset response in the seated condition may be, in part, explained differently. For example, given that participants always conduct the active condition before rewatching their movement in the seated condition, it seemed likely that there is a component of learning across the session that flow will sometimes be halted. This is confirmed with the analyses. The explanation that there is a visuomotor component here is given further weight by their conduction of an additional group of participants who perform the conditions in the reverse order, so this has strengthened the manuscript considerably. However, it does of course remain an imperfect control because the visual stimulus is now different between the conditions for these participants. It's the best that can be achieved with this type of paradigm though and of course it yields a great deal of ecological validity.

The reviewer is correct. But one should keep in mind that our result here stands in the context of a considerable amount of work on mouse cortex investigating responses to very similar visuomotor mismatches. There we can we have much additional evidence to argue that the cortical response to a visuomotor mismatch is a prediction error. We would argue, it is the best one can do in human experiments.

I was also wondering whether the authors may consider the findings in frontal electrodes more closely given that the title of the paper focuses on a specifically occipital effect. Their further analyses have confirmed that there are likely interesting frontal effects. From a theoretical point of view, the spatial dissociation in adaptation effects, which were stronger in frontal and weaker in occipital areas, seems interesting and perhaps worth discussing, especially given the interpretation that "mismatch processing may initially arise in sensory visual areas before engaging higher-order frontal regions." How come the frontal decrease in responses is not accompanied by an analogous decrease in its supposed occipital source? Could these two responses reflect different kinds of prediction error signals (i.e. objective vs subjective)?

We have added a paragraph to the Discussion addressing the differences between signals recorded over frontal and occipital electrodes, as suggested.

I remain concerned that the authors fight too defensively that they have absolutely isolated visuomotor prediction mechanisms with this paradigm. It's a nice, informative study, but it seems odd to argue there are no other possible explanations. One picks a design to optimize some features but they will always come at some cost to others. Prioritising ecological validity, which is a justifiable aim, necessarily usually weakens some control over confounds.

We are not sure what the reviewer is referring to here. We certainly do not think (or are aware of having argued) that a visuomotor prediction error is the only possibly interpretation of the response. In the last paragraph of our response to the reviewers point 3 in the last revision, we explicitly discuss that the interpretation of the responses as a prediction error is only one possible interpretation. Our argument is that it is the most likely given the evidence.

To outline my reasoning fully: My concerns wrt generic influences of action on perception are reflected in Fig 1. The P1 is smaller when walking than sitting. It seems likely that the mismatch response reflects something about extrapolation or prediction, because it is larger when walking. However, it's not necessarily sensorimotor prediction. Even if you remove action from the equation, the flow can be extrapolated or predicted most of the time in a way it cannot so well when the video is halted. Of course the sitting condition somewhat controls for it, but when it came second the visual flow disruptions were more predictable here. A reduction in effects over time is indeed confirmed with their analyses. They now have conducted a study with the conditions in the reverse order and they find the same thing. But of course this necessitates non-identical visual flow because the sitting condition is playing the previous participant's flow. So it is likely that across all of these comparisons, it is the visuomotor mismatch that is especially salient. It's just that each comparison is a bit messy/confounded. It would strengthen the manuscript if there were some consideration given to the other processes likely at play here.

We would be happy to add additional considerations to other processes. If the reviewer has anything specific in mind, we can add that, but it would need to be somewhat concrete with some theoretical basis. We share the reviewer’s intuition, but unless this can be formalized to the point of being experimentally testable, we do not see any value in discussing it in the manuscript.

Regarding the reason for a difference in visual responses in walking vs sitting state is, this is not entirely clear to us. Predictive processing would provide one possible explanation. Assuming the precision weighting of predictions is higher during walking, the sudden appearance of a visual stimulus might lead to stronger stimulus history prediction errors than when just sitting. But this is rather speculative.

As a more minor point in response to our previous review, whether particular accounts represent an 'orthodox' view at present does not determine whether they raise logical issues in need of consideration. The authors may have missed that the papers in question consider mechanisms underlying the attenuation of particular pieces of information ‘from perception’. Not perceptual processing. We have one percept at any one moment in time and must understand how different population types synergistically generate that percept.

Please excuse, the reviewer is correct, the orthodoxy of an idea is not relevant. For dubious reasons, we chose to euphemize what we actually meant to say here. With regards to circuit implementations of predictive processing (we cannot and do not intend to speak to interpretations of predictive processing that relate to conscious perception much of V1 activity is likely not consciously perceived – we assume this is what the reviewer is referring to by “we have one percept”) – the reviewers interpretation was not unorthodox, but rather incorrect (which is what we should have said). The statement that “the brain predictively ‘cancels’ expected action outcomes from perception” is incorrect in the context of sensory processing – based on both theoretical models of predictive processing, and more importantly physiological evidence. If the point was only in regards to conscious perception, we also suspect the statement is wrong, but even if it were correct, don’t see how it pertains to our work.

Similarly a little strange is the way in which the authors aggressively defend the position that self-generated motion is 'the strongest' type of prediction. Sure, we probably experience the effects of our actions more often than ambulances. But what about objects obeying laws of gravity or others' faces being structured and moving in systematic ways? It is hard to quantify, such that presumably many scientists would be skeptical of such a claim, and it is not needed logically to justify the importance of examining mechanisms enabling action to shape perceptual processing. I'd assume it better to fight the battles you need to (and can) fight, such that the robust claims carry more weight.

We believe it is absolutely essential for the progress of the field that we start to emphasize the differences between something that is “predictable in principle” and “predicted by the brain”. There is likely indeed a hierarchy of predictability that looks something like this:

(1) Sensorimotor coupling

(2) Laws of physics

(3) Behavior of other living things

(4) Artificial, human-made statistical relationships

Almost all published experiments are based on the fourth type of prediction. Indeed, why not use physics simulations instead of oddballs and MMN? We absolutely should! But the field tends to revert to artificial couplings. As a direct consequence of this, the number of papers appearing recently (from both human and mouse fields), that are built on the following premise:

(1) Expose an animal or human to an artificial coupling between A and B (e.g. an oddball, or a global oddball, or any of a myriad other constructions).

(2) Probe for prediction error responses to the violation of the artificial coupling.

(3) Find no prediction error responses and conclude predictive processing is wrong.

Is utterly baffling. The fallacy here is of course the assumption that if something is predictable in principle, the brain must predict it. Thus, we are, and will continue to be strong on this point, and we think it is essential that we – as a field – are.

Hope these comments are helpful.

Reviewer #2 (Public review):

Summary:

This study investigates whether visuomotor mismatch responses can be detected in humans. By adapting paradigms from rodent studies, the authors report EEG evidence of mismatch responses during visuomotor conditions and compare them to visual-only stimulation and mismatch responses in other modalities.

Strengths:

Authors use a creative experimental design to elicit visuomotor mismatch responses in humans.

The study provides an initial dataset and analytical framework that could support future research on human visuomotor prediction errors.

Weaknesses:

Methodological issues (e.g., volume conduction) make it difficult to confidently attribute the observed mismatch responses to activity in visual cortical regions. This could be alleviated by increasing the number of channels.

We have added a discussion of this.

The authors successfully demonstrate that visuomotor mismatch paradigms can, in principle, be applied in human EEG. This approach provides a translational bridge between rodent and human work on predictive processing.

Reviewer #3 (Public review):

Solyga, Zelechowski, and Keller present a concise report of an innovative study demonstrating clear visuomotor mismatch responses in ambulating humans, using a mobile EEG setup and virtual reality. Human subjects walked around a virtual corridor while EEGs were recorded. Occasionally, motion and visual flow were uncoupled, and this evoked a mismatch response that was strongest in occipitally placed electrodes and had a considerable signal to noise ratio. It was robust across participants and could not be explained by the visual stimulus alone.

This is an important extension of their prior work in mice, and represents an elegant translation of those previous findings to humans, where future work can inform theories of e.g. psychiatric diseases that are believed to involve disordered predictive processing. For the most part, the authors are appropriately circumspect in their interpretations and discussions of the implications. The paper in its current form represents an important addition to the literature.

The authors have included analyses of the auditory mismatch using temporal electrodes, referenced to Cz (and therefore should exhibit a mismatch positivity). This added data clearly and convincingly shows that the sensorimotor mismatch is, indeed, stronger than the passive auditory MMN.

The reference electrode placed at Cz makes it is difficult to interpret relative differences between frontal and occipital electrode responses, as the occipital electrodes are placed farther away from the Cz reference than the frontal electrodes. Similarly, signal occuring cortically near the Cz reference might only appear as though it is occipitally distributed in this montage. It is common in EEG research to remontage the data to an averaged common reference in order to better interpret the scalp distributions. As the electrode coverage was sparse for some subjects, this could be challenging, and this reviewer does not feel that it is necessary to do this analysis step, or even to drastically rewrite the body of the paper. We only request that some discussion, however brief, is included in the discussion section or the methods that recommend more dense electrode coverage in the future to better interpret scalp distributions and potential meso-scale sources.

We have added a discussion of this as suggested.

This is just a suggestion. The authors are encouraged to analyse (and report) time-frequency power and phase locking for these mismatch responses, as is common in much of the literature (see Roach et al 2008 Schizophrenia Bulletin). This is not to say that doing so will yield insights into oscillations per se, but converting the data to the time-frequency domain provides another perspective that has some advantages. fosters translations to rodent models, as ERP peaks do not map well between species, but e.g. delta-theta power does (see Lee et al 2018 Neuropsychopharmacology; Javitt et all 2018 Schizophrenia research; Gallimore et al 2023 Cereb Ctx). Further, ERP peaks can be influenced by the actual neuroanatomy of an individual (especially for quantifying V1 responses). Time frequency analyses may aid in interpreting the "early negative deflection with a peak latency of 48 ms " finding as well. As it stands, the report is complete, and it would be acceptable if the authors chose to save this type of analysis for a future publication.

We have added this as suggested.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

The authors have addressed most of my concerns by providing additional analyses, partly based on new data. The volume conduction issue is partly addressed based on the result showing latency differences, however to confidently assign responses to visual regions, one would need to perform recordings with a larger number of electrodes, sufficient to perform source localization. Nevertheless, the manuscript is now more solid than the previous version.

We have now added this point to the Discussion.

Reviewer #3 (Recommendations for the authors):

The reviewer appreciates that the authors have carried out time-frequency analyses, and are ok with them leaving this out of this paper.

We have now added this to the manuscript.

Finally, in response to the participant quote "are you printing this? hi mom!" - this reviewer concedes that it does not significantly detract from the report, and, in the interest of amusement and joy, would abide its reinstatement.

We greatly appreciate the reviewers entertaining our attempts at humor but will leave it out as originally suggested.

Version 3

Abstract

Likely the strongest predictor of visual feedback is self-motion. In mice, the coupling between movement and visual feedback is learned with first visual experience of the world (Attinger et al., 2017), and brief perturbations of the coupling result in strong visuomotor mismatch responses in visual cortex that possibly reflect prediction errors (Keller et al., 2012; Zmarz and Keller, 2016). In humans, predictive coding has primarily been studied using oddball paradigms which rely on violations of stimulus probability based on recent sensory history. It was still unclear, however, whether humans exhibit visuomotor mismatch responses similar to those observed in mice. This question was important for two reasons. First, visuomotor mismatch responses in humans constitute a basis to start translating the mechanistic understanding of the circuit that computes these responses from mouse to human cortex. Second, a paradigm that can trigger strong prediction error responses and consequently requires shorter recording times would simplify experiments in a clinical setting. Here, by combining a wireless EEG recording system with a virtual reality headset, we found robust visuomotor mismatch responses in human cortex that were characterized by a reversed polarity relative to visual evoked responses and a greater signal power than both visual responses and oddball mismatch responses.

Introduction

Predictive processing is a framework for understanding brain function. It proposes that the brain constructs internal models of the world based on regularities in past sensory input and sensorimotor loops to predict upcoming sensory input. The brain does this by minimizing the mismatches between predicted and actual sensory input. These mismatches, known as prediction errors, play a dual role in both updating internal representations and internal models (Friston, 2005; Keller and Sterzer, 2024; Rao and Ballard, 1999).

Predictions about sensory input can be made based on a variety of sources of other information the brain has available. Certain stimuli can be predicted from the preceding sensory input – experimentally this can be used, e.g., in oddball paradigms to trigger stimulus history based prediction errors (Garrido et al., 2009). Predictions can be crossmodal. Certain sounds are associated with specific types of visual input. Experimentally, visual stimuli can be coupled to sounds to then trigger audiovisual prediction errors in mouse visual cortex (Garner and Keller, 2022). Predictions can also be based on spatial memory. Not seeing a stimulus in a specific spatial location can result in visuospatial prediction errors in mouse cortex (Fiser et al., 2016). However, the strongest predictor - by frequency of experience, consistency in coupling, and precision in timing - is self-generated movement. Every movement is directly coupled to sensory feedback throughout life. Almost all visual and proprioceptive sensory input is the consequence of self-motion.

An essential ingredient to the computation of visuomotor prediction error responses are motor-related predictions of visual feedback (Keller and Mrsic-Flogel, 2018). In the mouse, evidence for such predictions has come from the discovery of a strong motor-related drive of activity in visual cortex (Keller et al., 2012; Niell and Stryker, 2010; Saleem et al., 2013). These motor-related signals are likely in part driven by motor-related predictions arriving from motor areas of cortex (Leinweber et al., 2017). However, the overall level of this motor-related activity is higher than one would expect simply from a motor-related prediction of visual input. Motor related activity is also evident in complete darkness and even prior to first visual experience in life (Attinger et al., 2017; Keller et al., 2012; Mahringer et al., 2022). It is possible, that this relatively high level of motor related activity in mouse visual cortex is a reflection of the fact that mice rely much less on fine visuomotor control.

In systems that have evolutionarily been driven to very high levels of precision of sensorimotor control, like vocal learning, there is much less such motor related activity. This is evident in the auditory system of a songbird, which relies on very temporally precise sensorimotor error detection for vocal learning, where there is much less motor related drive of auditory responses (Keller and Hahnloser, 2009). Similarly, there is very little movement related drive of activity in visual cortex of non-human primates (Liska et al., 2024; Talluri et al., 2023). In humans, movement-related modulation of EEG activity in visual cortex has been described (Cheng and Nordin, 2025; Gramann et al., 2010), but it is still unclear if visual responses are modulated by these signals. If indeed the evolutionary pressure for precision in visuomotor coupling determines the amount of motor modulation of visual responses, we would expect to find less of it in humans than in the mouse. Humans have a much higher reliance on visual feedback to guide movements than mice do.

This heightened reliance on visual guidance suggests that the human brain should be particularly adept at detecting mismatches in visuomotor coupling. Indeed, violations of the coupling between actual hand position and that of a virtual hand trigger responses in human visual cortex (Stanley and Miall, 2007). Even violations of simple action-outcome rules can trigger robust error responses in human subjects (Kimura and Takeda, 2019, 2014). Thus, we would also expect to find strong visuomotor mismatch responses in humans, comparable to – or stronger than – those observed in mice (Keller et al., 2012).

Results

Visual responses

To quantify visual responses in freely moving human participants, we combined a virtual reality (VR) headset with an 8-electrode wireless EEG recording system (Methods; Figure 1A). Participants were shown a virtual environment that consisted of a large empty floor space with a square checkerboard pattern fixed to their viewing angle directly in front of them (Figure 1B). The checkerboard covered a visual angle of 53° and contained a red fixation dot in the center. Participants were asked to fixate on the red dot throughout all measurements. The checkerboard reversed white-black at intervals sampled from a uniform distribution between 2 and 4 s (Figure 1C). Recordings were split into a passive session during which participants were seated on a chair, and an active session during which they were instructed to walk around a rectangular floor area of 5 by 7 m. During the entire experiment, we tracked the location of participants (Figure 1D).

EEG responses to visual stimulation are modulated by walking.

(A)Computer-generated image showing a participant wearing a wireless EEG recording setup and a virtual reality headset. An Open BCI EEG electrode cap is combined with META QUEST 3. Inset: First person view of the virtual environment. Red fixation dot was enlarged in figure to make it visible. (B)A third person view of the virtual scene used to study visual responses to a reversing checkerboard pattern. The black line on the ground indicates the safety boundary within which the participant was instructed to walk around during the movement block of the paradigm. The grid indicates 1 by 1 m squares. (C)Participants viewed a reversing checkerboard in two sessions: During half of the trials, they remained seated, and during the other half they walked within a defined safety boundary. Checkerboard reversed colors at random intervals every 2 to 4 s. (D)3D trajectory of an example participant during the visual experiment. Movement was recorded using accelerometers integrated into the VR headset. (E)Visually evoked potentials measured in the sitting session on the occipital electrodes O1 and O2. Responses recorded on other electrodes are presented in Figure S3. Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red line indicates the time of the checkerboard reversal. P1 - positive peak at around 88 ms. (F)As in E, but while participants were walking. Arrow indicates an early negative deflection preceding the P1 component. Inset: Comparison of visual evoked potentials measured in the sitting E and walking F sessions. The horizontal bar above the plot marks time bins in which responses differ significantly (black: p < 0.05) or do not differ (gray: p > 0.05).

Walking triggered movement related artifacts in the EEG signals (Figure S1). The strength of these movement artifacts varied as a function of hairstyle and gait of the participants. All participants were encouraged to go ‘gentle’ to reduce this problem. We excluded all data with eye blink and high movement related artifacts from further analysis (Figure S2, Methods; for number and percentage of excluded trials per session see Table S2).

We could detect multiphasic responses to the checkerboard inversion in occipital EEG electrodes (Figures 1E and S3). These responses were comparable to previous reports (Drislane, 2007) and exhibited a positive peak at around 88 ms (P1). Interestingly, when participants were walking, the same responses showed an early negative deflection with a peak latency of 48 ms that preceded P1 (Figure 1F). This early negative deflection at around 50 ms is also apparent in data from earlier studies comparing visual responses in walking and standing participants (Gramann et al., 2010).

While participants were instructed to fixate on the red dot, it is possible that there are systematic differences in eye movements between walking and sitting sessions. Our setup did not allow for concurrent eye tracking; thus, we cannot exclude the possibility that differences in eye movements contribute to the differences in EEG responses we observe. However, given the relatively rapid onset of response differences (40 ms), we suspect that stimulus-triggered eye movements cannot account for these early differences.

Visuomotor mismatch responses

To measure visuomotor mismatch responses, participants were instructed to walk around in a virtual corridor (Figure 2A). The corridor was 1 m wide, 2.4 m high and of an oval shape covering approximately 7 by 5 m (Figure 2B). Virtual movement in the corridor was coupled to the movement of participants. We refer to this as closed loop coupling. To trigger visuomotor mismatches, we briefly (0.5 s) halted the coupling between movement of the participants and visual feedback in the virtual corridor at random times (every 10 to 15 s; Methods; Figure 2C). Throughout all experiments, we tracked the virtual position of participants as they were walking around the corridor (Figure 2D).

Visuomotor mismatch elicits strong EEG responses.

(A)Computer-generated image showing a participant navigating a virtual tunnel. Inset: First person view of the tunnel. (B)Third person view of the virtual environment used to study visuomotor mismatch responses. The grid indicates 1 by 1 m squares. (C)In closed loop sessions, the walking speed of participants was coupled to movement in the virtual corridor. Visuomotor mismatches were introduced by briefly halting the visual flow for 0.5 s. Following each visuomotor mismatch event, the view was updated to the participants’ current position (brief peaks in visual flow speed after the mismatch event) and visuomotor coupling was resumed. (D)Trajectory of an example participant during the visuomotor mismatch paradigm. (E)Responses to visuomotor mismatches recorded from occipital electrodes (O1 and O2). Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset and offset of the visuomotor mismatch. The gray shaded areas mark the analysis windows used to quantify response strength in G and H. Analysis windows of 100 ms were centered on the peak of the visuomotor mismatch response and the teleport response. (F)As in E, but for visual flow playback halt responses recorded from occipital electrodes. (G)Comparison of the response strength to visuomotor mismatches and playback halts. Here and elsewhere: Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from one participant. ***: p<0.001. See Table S1 for all statistical information. (H)Comparison of the response strength to teleport events following visuomotor mismatches in the closed loop session and playback halts in the open loop session. n.s.: not significant. See Table S1 for all statistical information. (I)Average walking speed of participants during visuomotor mismatches. Inset: Temporally expanded view. The horizontal bar above the plot marks time bins where walking speed differs significantly (black: p < 0.05) or does not differ (gray: p > 0.05) from baseline.

Anecdotally, the perception some participants reported experiencing during these visuomotor mismatches was a very salient sudden movement of the entire visual world. In the words of one participant: “The world suddenly flew forward!”. This was reminiscent of a case report of a patient with a lesion to the lateral rectus muscle described by von Helmholtz (von Helmholtz, 1867): The lateral rectus muscle moves the eye laterally. The patient, upon attempting to move the affected eye laterally, reported seeing the world rapidly moving in the direction of the intended eye movement. One interpretation of this is that the combination of movement and absence of resulting visual flow change results in the perception that the world must be moving.

In the EEG recordings of occipital electrodes, we found strong responses triggered by these visuomotor mismatches (Figure 2E). Responses were dominated by a positive component peaking at 180 ms. Following the 0.5 s halting of the visual stimulus, the virtual location was updated to match the actual location of participants (we call this a teleport event), and the visual flow coupling was resumed. This resulted in a sudden change of visual stimulus combined with a visual flow onset and drove what is likely a visual response in the EEG signal (Figure 2E).

To quantify how much of the visuomotor mismatch response can be explained by visual input alone, we exposed participants to a replay of the visual flow that was self-generated in the preceding closed loop session, including the visual flow halts that constitute visuomotor mismatches in the closed loop session. We refer to this as open loop coupling. For these experiments, participants were seated on a chair. To reduce the likelihood of triggering nausea in participants, roll and pitch movements of the head were removed from this replay. We found measurable responses to visual flow playback halts (Figure 2F), but these were much smaller than visuomotor mismatch responses (Figure 2G). In the few participants who volunteered to experience playback that included pitch and roll movements, playback halt responses were also significantly smaller than visuomotor mismatch responses (Figure S4A-S4C) and not different from those in the reduced open loop session (Figure S4D and S4E). The difference in response amplitude between mismatch and playback response could not be explained by a simple movement related gain of visual responses, as the response to the teleport and re-onset of visual flow was similar in both closed and open loop sessions (Figure 2H). Note that the mismatch responses and playback halt responses shown here do not come from fully overlapping pools of participants (Table S1). A subset of visuomotor mismatch sessions had to be excluded because of recording noise, while some participants aborted the open loop recording due to developing nausea (Methods; for number and percentage of excluded trials per session see Table S2). A participant-wise comparison of mismatch and playback halt responses is shown in Figure S5.

By design, open loop sessions in the form of an identical playback (to match the visual statistics of open and closed loop sessions) always followed closed loop sessions, which could mean that the difference between visuomotor mismatch and playback halt responses could be driven by some form of adaptation. To control for this, we repeated the experiment in a separate cohort of participants with the order of sessions reversed: Participants first viewed a playback of the closed loop session of the previous participant and only then performed an active closed loop session. We found that independent of the order of closed and open loop sessions, visuomotor mismatch responses were larger than playback halt responses (Figure S6), thus excluding the possibility that the response difference may be driven by adaption or a reduction in salience of the sudden flow halt.

Although participants reduced their walking speed in response to the visuomotor mismatch when averaged over all trials, this did not reach statistical significance (Figure 2I). But see Figure 4A, for evidence that there was a significant reduction in walking speed when analyzing only the first few mismatch trials. In either case however, the reduction in walking speed was of much longer latency compared to the visuomotor mismatch responses, demonstrating that walking speed changes cannot explain visuomotor mismatch responses.

To test whether visuomotor mismatch events trigger eye movements that could explain visuomotor mismatch responses, we measured eye movements triggered by visuomotor mismatch in an independent cohort of participants. As with the EEG experiments, participants were exposed to a closed loop session with visuomotor mismatches. Throughout the experiment, we tracked eye position and eye blinks using an eye tracker built into the VR headset (Methods). Note, this VR headset was different from the one we used in the EEG experiments. We found that visuomotor mismatch triggered detectable eye responses. While we could not detect any systematic eye movements triggered by mismatch (Figure S7A and S7B), participants exhibited a reduction in average eye velocity (Figure S7C) and a reduction in eye blink rate (Figure S7D) with a latency of about 208 ms and 541 ms, respectively. Thus, both of these changes had latencies longer than those we observed for responses to visuomotor mismatch in the EEG experiments (Figure S7E). Thus, we find no evidence that visuomotor mismatch responses can be explained by eye blinks or eye movements.

Finally, we performed time-frequency power and phase-locking analyses of visuomotor mismatch and playback halt responses. The time-frequency power analysis revealed an increase in delta-band power during visuomotor mismatch, consistent with previous reports linking delta activity to prediction error processing, including reward prediction errors (Cavanagh, 2015), unexpected final words (Webb and Sohoglu, 2025), and visual deviance detection (West et al., 2024). Notably, it appears as if the increase in delta power emerged first over occipital electrodes and appeared later over more frontal electrodes, forming a spatiotemporal gradient of onset across the scalp (Figure S8). Delta power changes were markedly reduced in the playback halt responses and power changes occurred primarily at visual flow onset rather than at flow offset. The inter-trial phase coherence analysis revealed a strong delta-band synchronization over occipital electrodes following visuomotor mismatch, while the playback halt response primarily resulted in phase synchronization in both delta and theta bands following visual flow onset (Figure S9). These findings suggest that visuomotor mismatch evokes a delta-dominated low-frequency response that emerges first over occipital electrodes, then spreads across the cortex, and cannot be explained by passive visual flow changes alone. Unlike auditory oddball MMN responses, which show transient increases across a broader 3-12 Hz range (Ko et al., 2012), our response is more selectively delta-band dominant. This pattern more closely resembles visual deviance responses (West et al., 2024), where delta activity was stronger than theta and more prolonged, matching the temporal dynamics observed in our data.

Distribution of visuomotor mismatch response strength

In mice, visuomotor mismatch responses originate in primary visual cortex and propagate to a network of areas across dorsal cortex (Heindorf and Keller, 2023; Takeuchi et al., 2024). Thus, we expected to find larger and faster responses in the two occipital electrodes. To test this, we compared the visuomotor mismatch responses across the eight recorded locations in both response strength and timing (Figures 3A and 3B). We indeed found that the strongest responses were observed at occipital electrodes O1 and O2, though significant responses were also present at frontal (Fp1-2) and parietal (P3-4) sites (Figure 3C). We found a significant difference in response latency between occipital and frontal electrodes, with occipital signals occurring significantly earlier than frontal responses. (Figures 3D and 3E). This suggests that mismatch processing may initially arise in sensory visual areas before engaging higher-order frontal regions.

Visuomotor mismatch responses are most prominent over occipital cortex.

(A)Top down view of EEG electrode locations on the head. (B)Visuomotor mismatch responses measured on electrodes shown in A. Solid black lines represent the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (C)Comparison of the response strength to visuomotor mismatches measured on electrodes shown in A. Average response strength was calculated within a 100 ms window centered on the peak of the average visuomotor mismatch response across all electrodes. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from an individual participant. ***: p<0.001, **: p<0.01, *: p<0.05, n.s.: not significant. See Table S1 for all statistical information. (D)The responses shown in B, averaged over pairs of electrodes and overlaid. Solid black lines represent mean and shading SEM across electrodes. Dashed vertical red lines are onset and offset of the visuomotor mismatch. (E)Comparison of the latency to half maximum response for the 4 electrode pairs. **: p<0.01, n.s.: not significant. See Table S1 for all statistical information.

Temporal dynamics of neuronal and behavioral responses to visuomotor mismatch across a session

To examine how mismatch responses evolved over the session, we compared responses in early (first five) and late (last five) trials. We found that participants reduced their walking speed with a delay of about 500 ms following mismatch events. This effect was more prominent in early trials (14.3%) compared to late trials (5.7%) (Figure 4A). Interestingly, visuomotor mismatch responses mirrored this pattern over frontal electrodes with a clear attenuation from early to late trials (Figure 4B). By contrast, response changes in occipital electrodes were smaller between early and late trials (Figure 4C). The fact that the reduction over frontal cortex is larger than that of responses over visual cortex (Figure 4D), indicates that there may be a gradient of experience dependent changes in processing from early sensory to frontal areas of cortex that relate to changes in behavioral responses to, but not detection of prediction errors.

Visuomotor mismatch responses show experience-related changes predominantly over frontal electrodes.

(A)Average walking speed during visuomotor mismatch events in early versus late trials. The solid black line shows the mean and the shaded area the SEM across participants. Vertical red dashed lines indicate mismatch onset and offset. The horizontal bar above the plot marks time bins in which the walking speed in early trials differs significantly (black, p < 0.05) or not (gray, p > 0.05) from that in late trials. (B)Visuomotor mismatch responses in early versus late trials over frontal electrodes. (C)As in B, but for occipital electrodes. (D)Comparison of the difference between visuomotor mismatch response in early and late trials for frontal and occipital electrodes. Average response strength was calculated within a 100 ms window centered on the peak of the average visuomotor mismatch response in early trials above frontal or occipital electrodes. Boxes mark median, quartiles, and range of data not considered outliers. *: p<0.05. See Table S1 for all statistical information.

Comparison of visual responses and visuomotor mismatch responses

The time course of visual responses (Figure 1E) and visuomotor mismatch responses (Figure 2E) differed in that they appeared to exhibit reversed polarity (Figure 5A). Despite this difference in polarity, the initial peaks occurred at similar latencies. The total power of the visuomotor response was higher than that of the visual response (Figure 5B), but given that we did not optimize either the visuomotor mismatch stimulus nor the visual stimulus for maximum response, this comparison is more qualitative than quantitative in nature. We include it here primarily to illustrate how the visuomotor mismatch response compares to a more conventional visual response.

Visuomotor mismatch responses have reversed polarity and more power compared to visual responses.

(A)Comparison of visual (Figure 1E) and visuomotor mismatch (Figure 2E) responses recorded from all electrodes. Solid black line represents the mean, and shading indicates the SEM across participants. Dashed vertical red lines are onset (visual and mismatch) and offset (mismatch) of the stimuli. (B)Comparison of the power of visual and visuomotor mismatch responses, calculated within a 0 - 0.5 s time window following stimulus onset. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from an individual participant. ***: p<0.001, *: p<0.05. See Table S1 for all statistical information.

Comparison of visuomotor mismatch responses and auditory oddball mismatch responses

To further contrast the strength of visuomotor mismatch to other frequently used prediction error signals, we also recorded EEG responses in an auditory oddball paradigm. Mismatch responses recorded with this paradigm can be thought of as a stimulus history prediction error (Garrido et al., 2009; Näätänen et al., 2007). We implemented an oddball paradigm composed of a sequence of identical tones that was occasionally disrupted by a tone of a different frequency (Figure 6A). We used pure tones of 1 kHz and 1.2 kHz frequency and alternated in blocks which tone was standard and which was the deviant (Methods). Participants watched short silent movies in the VR headset while listening to the tone sequences delivered in headphones. To isolate the oddball driven mismatch response, we subtracted the response to the tone when presented as a standard from the response to the same tone when presented as a deviant. The resulting difference waveform closely resembled those reported in previous studies (Figures 6B-6C, S10, and S11) (Näätänen et al., 2007). Comparing these oddball mismatch responses to the visuomotor mismatch response in occipital electrodes, we found that both had similar dynamics with a dominant positive peak at around 180 ms (Figure 6D). Comparing the total power of the two responses, we found that visuomotor mismatch responses were significantly larger than oddball mismatch responses (Figure 6E). This was also true for responses measured in temporal electrodes (Figures S10 and S11). But also here, the comparison of total power is more qualitative than quantitative as neither stimulus was optimized for maximum EEG response, and we include it to illustrate how typical variants of these responses compare.

Visuomotor mismatch responses are larger than auditory oddball mismatch responses but have similar temporal dynamics

(A)Design of the auditory oddball paradigm with examples of the silent films participants were exposed to. (B)Top: Auditory responses to the 1 kHz tone presented as a standard versus as a deviant. Bottom: Oddball mismatch response calculated by subtracting the average response over trials in which the tone was presented as a standard from those when the tone was presented as a deviant. Solid black lines represent the mean, and shading indicates the SEM across participants. Dashed vertical red line is the onset of the auditory stimulus. (C)As in B, but for the 1.2 kHz tone. (D)Comparison of visuomotor mismatch, oddball mismatch (average over data shown in panels B and C) and playback halt responses recorded from occipital electrodes. Solid black lines represent the mean response, with shading indicating SEM across electrodes. Dashed vertical red lines are onset (visual, mismatch) and offset (mismatch) of the stimuli. (E)Comparison of the power of visuomotor mismatch, oddball mismatch response (average over data shown in panels B and C) and playback halt responses, calculated within a 0 s - 0.5 s time window following stimulus onset. Boxes mark median, quartiles, and range of data not considered outliers. Each circle represents data from an individual participant. ***: p<0.001; *p<0.05. See Table S1 for all statistical information.

Discussion

The utility of paradigms for measuring prediction errors in humans primarily comes from their relevance to psychiatric disorders, where disruptions in prediction error processing are widely implicated (Adams et al., 2013; Fletcher and Frith, 2009; Kirihara et al., 2020; Qela et al., 2025; Sterzer et al., 2018). Various types of prediction error responses are being explored as potential biomarkers in clinical research, serving to monitor disease progression, evaluate symptoms, and provide insight into the underlying neuronal mechanisms of these conditions.

The most prominent of these prediction errors is mismatch negativity (MMN). MMN is the first component of the oddball mismatch response (Näätänen et al., 1978). MMN has been shown to be reduced or abnormal in patients with schizophrenia spectrum disorders (Todd et al., 2012; Umbricht and Krljes, 2005), autism spectrum disorders (Chen et al., 2020; Dunn et al., 2008; Schwartz et al., 2018), speech disorders (El Hatal de Souza et al., 2020), depression (Tseng et al., 2021), and bipolar disorders (Chitty et al., 2013). More generally, it is thought that many of these disorders can be related to changes in predictive processing. In the case of schizophrenia, for example, the symptoms are thought to arise from an imbalance in the strength of high and low level predictions (priors) (Schmack et al., 2017, 2015a, 2015b, 2013). Further testing these ideas will require experiments that trigger identified functional responses in humans, for which we have a circuit level understanding based on cell type specific recordings.

We were able to measure robust visuomotor mismatch responses in freely moving humans (Figure 2). These signals cannot be attributed to changes in participants’ behavior or to simple visual offset responses. Notably, the visuomotor mismatch responses exhibited a markedly different temporal profile compared to purely visual responses (Figure 5). What might account for this difference? One possibility is that visuomotor mismatch signals may rely more strongly on input from other cortical areas than visual responses and are thus delayed relative to visual responses. A more interesting interpretation is that visual responses and visuomotor mismatch responses are both prediction errors of different types. A central tenet of the cortical circuit for predictive processing is the split into separate populations of neurons that compute positive and negative prediction errors (Keller and Mrsic-Flogel, 2018; Rao and Ballard, 1999). In this interpretation, a visuomotor mismatch response is a negative prediction error, while the response to a visual stimulus is a positive prediction error.

Visuomotor prediction errors are likely computed in layer 2/3 of mouse visual cortex (Jordan and Keller, 2020), and are also preferentially detectable in superficial layers of human visual cortex (Thomas et al., 2024). In the mouse, positive and negative prediction errors are likely computed in separate populations of neurons in visual cortex (Keller and Mrsic-Flogel, 2018; O’Toole et al., 2023). These two cell types have a different depth distribution, with negative prediction error neurons more superficial and positive prediction error neurons located deeper in cortex (O’Toole et al., 2023). Given that EEG signals are thought to exhibit positive or negative deflections as a function of depth of the source (Cohen, 2017; Kirschstein and Köhling, 2009), the polarity reversal of visual and visuomotor mismatch responses (Figure 5) may be the consequence of different populations of cells being activated and inhibited. Visuomotor mismatch should activate negative prediction error neurons and inhibit positive prediction error neurons, while the reverse is the case for a visual stimulus.

The fact that there is also a strong visual response to the visual flow re-onset following visuomotor mismatch means that our visuomotor mismatch paradigm might allow us to measure both negative and positive prediction error responses. More intriguingly, there is a differential modulation by walking of the two responses (Figures 2G and 2H). In theory, visuomotor mismatch drives a combination of two negative prediction errors. One is based on a movement related prediction. The other on a stimulus-history related prediction. Both the self-motion as well as the ongoing visual flow would predict a continuation of visual flow. A visuomotor mismatch violates both predictions. During passive observation, there is no movement-related prediction, but the stimulus history-based prediction is still violated. This could explain the strong reduction of the negative prediction error response during sitting, and may also explain why there is no reduction for the positive prediction error response.

Visuomotor mismatch responses recorded over frontal electrodes adapted more strongly with experience than responses recorded over occipital electrodes (Figure 4). This may suggest that sensory areas primarily act as detectors of visuomotor mismatch events (subjective prediction error signal), whereas frontal areas reflect higher-level adaptation to the event and the resulting behavioral adjustment. Mouse data suggested that visuomotor mismatch responses originate in visual sensory areas (Heindorf and Keller, 2023; Jordan and Keller, 2020) and are broadcast to other cortical areas from there. Thus, the stronger decrease over frontal cortex suggests that signals measured over frontal electrodes may reflect the integration of sensory mismatch information with behavioral or attentional signals. This behavioral component makes the paradigm particularly interesting as a potential clinical biomarker, as it may allow us to assess how quickly participants adapt their behavior to mismatch event and how this adaptation relates to prediction error responses in frontal cortex.

Intriguing is the similarity in timing and polarity of the playback halt and oddball response (Figure 6D). An oddball response is always a combination of both a positive and a negative prediction error. There is a negative prediction error for the absence of the expected standard tone as well as a positive prediction error for the presence of the unexpected oddball tone. Thus, it is conceivable that the subtraction of standard response from the oddball response, isolates the component of the oddball response that corresponds to a negative prediction error.

One of the key challenges in systems neuroscience is translating findings from animal models to humans. Although recent animal studies have provided detailed insights into the circuit level implementation of predictive processing in the cortex (Keller and Mrsic-Flogel, 2018), and psychological and psychiatric conditions have long been described within this framework (Keller and Sterzer, 2024; Sterzer et al., 2018), translation to human research has been limited by a lack of methods to record cell type specific signals in human experiments. With an understanding from animal models of how positive and negative prediction errors are computed, and which cortical layers are involved, an approach based on functionally identified responses might be most promising. One step in this direction is to use experimental paradigms that can cleanly separate e.g. positive and negative prediction error responses. As argued above, we speculate that the visuomotor mismatch paradigm as we have used it here, is one possible way to achieve this.

It should also be highlighted that the paradigm we have developed here is still limited in its clinical applicability by practical constraints, including movement-related artifacts, signal-to-noise levels, low channel density, and the requirement for active engagement. Nevertheless, active task engagement may itself be a critical feature for revealing deficits in cortical processing that remain undetectable in more passive paradigms. Thus, although active, more video-game-like paradigms may appear less practical in certain clinical contexts, they may be essential for capturing meaningful and behaviorally relevant neuronal responses. It will also be of importance to overcome a challenge of performing recordings with a higher channel density in freely moving participants for improved source localization. Thus, methodological refinement and hardware optimization will be required before reliable clinical translation. However, these limitations do not detract from the paradigm’s immediate utility as a research tool for studying prediction error processing.

Methods

Participants

The study was approved by Ethikkommission Nordwestund Zentralschweiz (Project-ID: 2024-02458). Participants signed a written informed consent before participation and were not financially compensated for their participation. A total of 91 recording sessions from healthy adults were included in the study. Sessions were obtained from participants aged 18–65 years, with a small number of participants contributing more than one recording session. The age distribution across sessions was as follows: 38 sessions from ages 18–30, 37 sessions from ages 31–40, 11 sessions from ages 41–50, and 5 sessions from ages 51–65. None of the participants reported a prior diagnosis of movement disorders, vestibular dysfunction, or epilepsy. Experience with virtual reality technology ranged from beginner to advanced, with most participants reporting minimal prior experience.

EEG recordings in humans

We integrated a wireless EEG recording system with a VR headset. The EEG recording system was composed of a wet electrode cap and a Cyton biosensing board from OpenBCI. This allowed us to record 8 EEG channels (43 sessions: Fp1, Fp2, C3, C4, P3, P4, O1 and O2; 48 sessions: Fp1, Fp2, T3, T4, T5, T6, O1 and O2) at a 250 Hz sampling rate. Electrode labels follow the international 10-20 system: Fp = frontopolar, C = central, P = parietal, O = occipital, and T = temporal; odd numbers indicate the left hemisphere, and even numbers the right (Figure 3A). All EEG data were wirelessly (via Bluetooth) transmitted to a nearby computer. The VR headset was a Meta Quest 3, with a virtual environment developed in the Unity engine (Unity Technologies). Virtual 3D objects were designed in Fusion 360 (Autodesk). Synchronization between EEG recording and VR headset was performed by connecting an auditory output of the VR headset to the Cyton board to exchange synchronization triggers. This allowed us to synchronize EEG data with VR events offline. Auxiliary signals, including the participant’s position, stimulus trigger timing, and type, were recorded directly on the headset at a sampling rate of approximately 100 Hz.

Eye movement measurements

We recorded eye blinks and eye movements using a Meta Quest Pro VR headset and its built in eye tracker during closed loop sessions. Eye-tracking data were sampled at 72 Hz. We combined blink signals from left and right eye, by averaging the two channels at each time point, yielding a single blink/eye-closure trace per participant. To compute eye movement speed, we first computed eye-movement speed separately for the left and right eye from the moment-to-moment change in eye rotation between consecutive samples and then averaged the two speed traces to obtain a single eye-speed measure.

Visual responses

Visual stimuli were presented using the VR headset. We used a reversing square checkerboard stimulus to drive visually evoked potentials. The checkerboard reversed colors at random intervals (between 2 and 4 s). In virtual space, the checkerboard measured 0.5 by 0.5 m and was positioned 0.5 m in front of participants’ eyes. This resulted in a horizontal and vertical coverage by the checkerboard of approximately 53° of visual angle. Each visual stimulation session lasted 4 min, during which the checkerboard reversed colors between 78 and 82 times. For half of the session, participants viewed the stimulus while seated; for the other half, they were instructed to walk freely within a 7 by 5 m empty floor space. The order of sitting and walking sessions was randomized across participants.

Visuomotor mismatch responses

For the experiments measuring visuomotor mismatch responses, we used a 3D virtual corridor with vertical gratings on the walls. The corridor measured 1 m in width, 2.4 m in height, and had an oval shape of 7 by 5 m. Visuomotor mismatches were introduced at random intervals every 10 to 15 s as participants walked through the corridor. The session lasted 5 min and included between 22 and 26 visuomotor mismatch events. During these events, the coupling between the participant’s movement and the visual feedback in the VR headset was briefly interrupted, i.e. the visual scene was frozen for 0.5 s. Because participants continued to move during this time, they were teleported to their current position in the virtual space following visuomotor mismatch event.

Playback halt responses

To quantify how much of the visuomotor mismatch response could be explained by visual input alone, participants were asked to passively observe a replay of the visual flow they had self-generated during the preceding closed loop session; we refer to this replay condition as the open loop session. We quickly learned, however, that watching 5 minutes of playback in the VR headset triggered nausea in most participants. Thus, we started experimenting with changes to the playback to minimize the risk of triggering nausea. One such modification we settled on to use for experiments was a playback version that omitted pitch and roll movements of the head. Full playback involved six degrees of freedom (6DOF): 3D position in space, plus pitch, yaw, and roll angles of the head. The constrained playback included only four degrees of freedom (4DOF): The 3D position in space, but only yaw movements of the head. A subset of participants volunteered to view the full playback (Figure S4). Open loop sessions were conducted while participants were seated and lasted 5 min, matching the duration of the closed loop session. Approximately 23% of participants reported strong nausea during the beginning of the 4DOF open loop session, at which point the recordings were terminated.

Mismatch negativity

To measure mismatch negativity, we used an auditory oddball paradigm. Auditory stimuli consisted of two pure tones: 1 kHz and 1.2 kHz. They were presented at 60 to 70 dB sound pressure level in a counterbalanced design, in which each tone served as the standard in one condition and as the deviant in the other. Tones were 50 ms in duration, including 5 ms linear rise and fall ramps. Stimuli were delivered in a pseudorandom order, with deviant tones comprising 10% of all trials and preceded by at least five standard tones. The inter-trial interval was selected from a uniform distribution of between 500 and 600 ms. Each session lasted 3 min and included between 28 - 33 deviant and 244 - 299 standard stimulus presentations. Participants were seated and listened to the tones while watching a silent movie that was not related to the tone sequences.

EEG Signal Analysis

Data analysis was done using custom-written MATLAB scripts. EEG signals were band-pass filtered between 1 and 100 Hz. To remove power line noise, a band-stop filter was applied between 40 and 60 Hz. Movement of the participants triggered all varieties of movement related artifacts in the EEG recordings. The strength of these artifacts depended on a variety of factors: Impedance of the electrodes, hair style of participants, gait pattern, and likely others. To reduce data contaminated by excessive movement artifacts and eye blinks, trials with a maximum absolute response amplitude exceeding 100 µV within the time window −0.5 to 1 s relative to the trigger were discarded from further analysis (Figure S2). Data from each electrode were included in the final analysis only if at least 15 triggers remained after exclusion of triggers with excessive movement artifacts (Table S1). To compare average response strength across sessions, a 100 ms analysis window was used, centered on the peak of the respective responses: The visuomotor mismatch event recorded at occipital electrodes O1 and O2 (Figures 2G, S4C, S5C, S6C), the visual flow re-onset response (Figure 2H), the visuomotor mismatch response in early trials for frontal and occipital electrodes (Figure 4D), the mean visuomotor mismatch response across all electrodes (Figure 3C), the mean visual response across all electrodes (Figure S3C) or playback halt response in the 6DOF session (Figure S4E). Signal power was compared by calculating the mean squared amplitude within a 0 - 0.5 s analysis window following stimulus onset (Figures 5B, 6E, S10E, S11D). To analyze how responses evolved over the course of the session, we computed moving-average responses across 10 trials at successive percentages of progress through the paradigm (Figure S6D and S6E). Time–frequency power and inter-trial phase coherence (ITPC) were computed using complex Morlet wavelet convolution approach (Cohen, 2014). EEG data were sampled at 250 Hz and analyzed from −0.5 to 2 s relative to the event of interest. Wavelets were generated from −1 to 1 s and covered 30 logarithmically spaced frequencies between 2 and 40 Hz. The number of cycles increased logarithmically with frequency, from 3 to 10 cycles.

Statistical tests

All statistical analyses were conducted using hierarchical bootstrap (Saravanan et al., 2020). Bootstrap resampling enables statistical comparisons across sessions without assuming a specific distribution of the EEG data. For analysis in Figures 1, 2, 4, 6, S4, S5 and S6, we averaged signals across electrode pairs (O1-2 or Fp1-2) and treated the result as a single data point per participant. For Figures S10 and S11, the same procedure was applied to temporal electrode pairs (T3-4 and T5-6). Likewise, in Figures 3D, 3E, and 5, we averaged signals from the respective electrode pairs O1-2, C3-4, P3-4, and Fp1-2, and treated each pair as a single data point per participant. For the analysis shown in Figures 3C and S3C we included signals from all electrodes and used a nested bootstrap to account for multiple data points originating from the same participant. We first resampled the data with replacement at the level of participants, followed by resampling at the level of electrodes. For each resampled population, we computed the mean response and repeated this procedure 10000 times. The p value was estimated as the fraction of bootstrap samples in which the sample mean violated the tested hypothesis. See Table S1 for all information on number of participants or electrodes used for all analyses shown.